---
title: "Running the FishKelp model for the Blamey and Bolton fish"
subtitle: "Instructions and results for Zoë"
author: "AJ Smit"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: journal
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
---

# Zoë, start here

I have prepared the FishKelp data for every fish listed in Table 2 of
Blamey and Bolton (2018). This file shows you what is available, how the model
uses it, and how to repeat the full analysis on your computer.

The paper lists 30 fish associated with South African kelp forest and temperate
reef ecosystems. That wording needs care. The table combines two habitats, and
its columns describe fisheries and exploitation status. I have used the table
to define the fish set, but not as evidence that each fish uses kelp. The
FishKelp literature assessment supplies that evidence.

Please keep the trophic and habitat routes separate. A fish may feed on kelp or
kelp-associated prey without using kelp structure as habitat. It may also occur
within a kelp forest without evidence of either action. The model retains those
distinctions.

# What I have packaged for you

The `data/fishkelp_v7` directory contains the published species list, the
FishKelp evidence records for those species, their traits, and the posterior
summaries from the production model. I have included the full model-ready R
object as well. You do not need access to my FishKelp directory to run this
analysis.

Five of the 30 fish do not have a posterior estimate in FishKelp v7.
*Chrysoblephus anglicus* falls outside the review boundary. *Coryphaena
hippurus* and *Polysteganus praeorbitalis* do not occur in the v7 review species
frame. *Seriola lalandi* and *Thyrsites atun* are in scope, but no model-ready
records occur for them in this snapshot.

Do not replace those missing estimates with zero. Missing means that the model
did not estimate the species. A zero would state that the evidence supports no
kelp relationship, which is a different claim.

```{r setup}
qmd_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
repo_root <- dirname(dirname(qmd_path))
bundle_dir <- file.path(repo_root, "data", "fishkelp_v7")

fish <- read.csv(
  file.path(bundle_dir, "blamey_bolton_2018_fish.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
packaged_estimates <- read.csv(
  file.path(bundle_dir, "fishkelp_posterior_estimates_v7.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  na.strings = c("NA", "")
)
estimates <- packaged_estimates[packaged_estimates$modelled_v7,]
estimates$species <- estimates$accepted_name
estimates <- estimates[order(-estimates$T_reach),]
```

# Check the species coverage

The status table below records all 30 fish. The record counts refer to evidence
from the full FishKelp literature assessment, not only to Blamey and Bolton
(2018).

```{r coverage-table}
coverage <- fish[, c(
  "table_order",
  "common_name_published",
  "scientific_name_published",
  "accepted_name",
  "model_status_v7",
  "n_model_records_v7",
  "n_analysis_records_v7"
)]
names(coverage) <- c(
  "Order",
  "Common name",
  "Published name",
  "FishKelp name",
  "Model status",
  "Model records",
  "Analysis records"
)
knitr::kable(coverage, caption = "Coverage of the 30 fish in Table 2.")
```

# Current model result

This report uses the production summaries packaged from FishKelp v7. The model
estimates 25 of the paper's fish.

```{r posterior-table}
posterior_table <- estimates[, c(
  "species",
  "T_reach",
  "T_reach_lo",
  "T_reach_hi",
  "H_reach",
  "H_reach_lo",
  "H_reach_hi"
)]
names(posterior_table) <- c(
  "Species",
  "T mean",
  "T 2.5%",
  "T 97.5%",
  "H mean",
  "H 2.5%",
  "H 97.5%"
)
knitr::kable(
  posterior_table,
  digits = 3,
  caption = "Posterior expected evidentiary reach for the 25 modelled fish."
)
```

```{r forest-plot}
#| fig-width: 10
#| fig-height: 11
#| fig-cap: "Trophic and habitat evidentiary reach for the 25 modelled fish. Points show posterior means and lines show 95% credible intervals."
plot_data <- rbind(
  data.frame(
    species = estimates$species,
    Route = "Trophic",
    Mean = estimates$T_reach,
    Lower = estimates$T_reach_lo,
    Upper = estimates$T_reach_hi
  ),
  data.frame(
    species = estimates$species,
    Route = "Habitat",
    Mean = estimates$H_reach,
    Lower = estimates$H_reach_lo,
    Upper = estimates$H_reach_hi
  )
)
plot_data$species <- factor(
  plot_data$species,
  levels = rev(estimates$species)
)

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = Mean, y = species, colour = Route)
) +
  ggplot2::geom_vline(
    xintercept = 0:4,
    colour = "grey90",
    linewidth = 0.35
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = Lower, xend = Upper, yend = species),
    linewidth = 0.55
  ) +
  ggplot2::geom_point(size = 1.8) +
  ggplot2::facet_wrap(~Route, nrow = 1) +
  ggplot2::scale_colour_manual(
    values = c(Trophic = "#A64833", Habitat = "#276678"),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    breaks = 0:4,
    labels = c("None", "Co-occurrence", "Use", "Selection", "Contribution"),
    limits = c(0, 4)
  ) +
  ggplot2::labs(
    x = "Expected evidentiary level",
    y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold"),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )
```

*Pachymetopon blochii* has the highest trophic reach among the fish. Its
habitat estimate is also high relative to the other modelled fish, although
*Poroderma africanum* has the highest habitat mean. These are relative evidence
positions. They are not probabilities that either species depends on kelp.

# Run the full model

Please use the production run when you need to regenerate the posteriors rather
than inspect the packaged results. Open a terminal at the repository root and
run:

```{bash, eval=FALSE}
Rscript R/run_fishkelp_v7_suite.R
```

The run uses four chains of 60,000 iterations, with the first 15,000 iterations
of each chain discarded. It retains 180,000 draws. On my current machine this
takes several minutes and uses appreciably more memory than rendering this
document.

The script fits the complete FishKelp dataset, which contains 3,648 records for
62 model species. It then extracts the 25 Blamey and Bolton fish with estimates.
Please do not subset the input before fitting. The species share model
coefficients and a correlated species-effect distribution, so an early subset
would change the estimates you are trying to reproduce.

The run checks every free fixed effect and cut-point. It stops when maximum
split-Rhat exceeds 1.05. It also compares the new species summaries with the
packaged FishKelp results. The largest difference must be no more than 0.01.

The new results are written to `model_outputs`. That directory is excluded from
Git because the posterior-draw file can be large. When the run finishes, render
this document again:

```{bash, eval=FALSE}
quarto render Bayesian_framework/FishKelp_full_species_suite.qmd
```

The new posterior summaries will be in
`model_outputs/blamey_bolton_fish_posterior_summary_v7.csv`.

# Reproduce the focal prior and posterior figure

The detailed *P. blochii* report includes the blue dashed prior distribution
and red posterior distribution that you wanted. Render it with:

```{bash, eval=FALSE}
quarto render Bayesian_framework/FishKelp_strongest_fish_posteriors.qmd
```

That document uses the same packaged data. Its retained draws save you from
repeating the full fit whenever you change the prose or figure formatting.

# Interpret the evidentiary scale

The horizontal scale represents the expected position in the evidence hierarchy,
namely None, Co-occurrence, Use, Selection, and Contribution. Dependence remains
level 5 in the review definitions, but no qualifying Dependence record occurs in
the fitted data.

Do not describe the expected level as a percentage. A posterior mean between
None and Co-occurrence expresses uncertainty across ordered evidence
categories. It does not mean that a fish is a given percentage dependent on
kelp. The ecological claim must remain tied to the source evidence and its
route.

# Updating the FishKelp data

I can refresh this package from a newer FishKelp checkout with:

```{bash, eval=FALSE}
Rscript R/package_fishkelp_v7_data.R /path/to/FishKelp
```

The package records the FishKelp commit and source checksums in
`data/fishkelp_v7/SOURCE_PROVENANCE.csv`. Please tell me before replacing the
snapshot, since a change to the evidence data requires a new model fit and a
check of every reported result.
