---
title: "Posterior distributions for the strongest kelp-associated fish"
subtitle: "A reproducible link to the FishKelp v7 evidence-hierarchy model"
author: "AJ Smit"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: journal
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
---

# Purpose

This document identifies the fish with the strongest modelled relationship with
kelp in the FishKelp v7 analysis. It then plots that species' trophic and habitat
posterior distributions. The two routes remain separate throughout.

The document does **not** convert the evidence hierarchy into a percentage of
"kelp dependence". The FishKelp response is ordinal: None, Co-occurrence, Use,
Selection, Contribution and Dependence. The fitted v7 data contain observations
from None to Contribution; Dependence is defined by the review instrument but is
not represented by a qualifying record. Consequently, the posterior quantity
shown here is the continuous **expected evidence level** on the observed 0--4
scale.

# Does the proposed framework make sense?

The ecological structure in `Bayesian_framework_Read_Me.qmd` is useful. In
particular:

- trophic and habitat evidence should be analysed separately;
- uncertainty and the specificity of each record should affect inference; and
- the ordered claims from co-occurrence through dependence should not be treated
  as synonyms.

The code in `Posteriors.qmd` is not yet a valid implementation of that idea. It
draws simulated observations from a selected prior and then fits a Beta density
back to those same observations. This is circular. The optimisation step does
not combine a stated prior with a data likelihood, does not use the FishKelp
records, and does not use Stan despite loading `rstan`. It also optimises the two
Beta shape parameters without constraining them to be positive.

A Beta distribution would be appropriate for an unknown continuous proportion
with a compatible sampling model. The FishKelp response is instead an ordered
evidence category. Assigning a high Beta prior merely because a fish occurs in
the west or at kelp depth also risks treating access to kelp as evidence of use
or dependence. That contradicts the evidence hierarchy. In the v7 analysis,
region and species traits are predictors, while the highest level supported by
each source is the response.

# Analysis inputs

The repository contains a dated copy of the model input, posterior summary,
evidence records and posterior draws used here. The paths below are relative to
this document, so the report will work from a fresh clone without access to
my FishKelp directory.

```{r setup}
qmd_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
analysis_dir <- dirname(qmd_path)
repo_root <- dirname(analysis_dir)
bundle_dir <- file.path(repo_root, "data", "fishkelp_v7")

model_data_path <- file.path(
  bundle_dir, "model", "tdr_model_data_v7.rds"
)
estimate_path <- file.path(
  bundle_dir, "fishkelp_posterior_estimates_v7.csv"
)
evidence_path <- file.path(
  bundle_dir, "fishkelp_model_records_v7.csv"
)

output_dir <- file.path(analysis_dir, "posterior_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

estimates <- read.csv(
  estimate_path,
  stringsAsFactors = FALSE,
  na.strings = c("NA", "")
)
estimates$species <- estimates$accepted_name
fish_estimates <- estimates[
  estimates$modelled_v7 &
    estimates$n_T > 0 & estimates$n_H > 0,
]
fish_estimates$mean_route_reach <- rowMeans(
  fish_estimates[, c("T_reach", "H_reach")]
)

# Match the ordering used in the FishKelp summary.
fish_estimates <- fish_estimates[
  order(-fish_estimates$T_reach, -fish_estimates$H_reach),
]
selected_species <- fish_estimates$species[1]
species_slug <- gsub("[^A-Za-z0-9]+", "_", selected_species)

top_habitat_fish <- fish_estimates$species[which.max(fish_estimates$H_reach)]
```

## Selection result

The FishKelp summary ranks evidence-bearing fish by trophic reach. That rule
selects *`r selected_species`*. It is also the fish with the highest unweighted
mean across the two routes. The habitat-only leader is *`r top_habitat_fish`*,
so *`r selected_species`* should not be described as the highest-ranked fish
under every possible definition of "strongest".

```{r ranking-table}
ranking_table <- head(
  fish_estimates[, c(
    "species", "n_T", "T_reach", "T_reach_lo", "T_reach_hi",
    "n_H", "H_reach", "H_reach_lo", "H_reach_hi", "mean_route_reach"
  )],
  10
)
names(ranking_table) <- c(
  "Species", "T records", "T mean", "T 2.5%", "T 97.5%",
  "H records", "H mean", "H 2.5%", "H 97.5%", "Mean of routes"
)
knitr::kable(
  ranking_table,
  digits = 3,
  caption = "The ten fish with the highest trophic expected reach, matching the FishKelp summary order."
)
```

The coded evidence for *`r selected_species`* has the following distribution
across observed levels. These are input records, not posterior probabilities.

```{r evidence-table}
evidence <- read.csv(evidence_path, stringsAsFactors = FALSE, check.names = FALSE)
names(evidence)[8] <- "level"
selected_evidence <- evidence[evidence$species == selected_species,]
evidence_counts <- as.data.frame(
  table(
    Route = factor(selected_evidence$route, levels = c("T", "H")),
    Level = factor(selected_evidence$level, levels = 0:4)
  )
)
evidence_counts <- evidence_counts[evidence_counts$Freq > 0,]
evidence_counts$Route <- c(T = "Trophic", H = "Habitat")[
  as.character(evidence_counts$Route)
]
evidence_counts$Level <- c(
  "0: None", "1: Co-occurrence", "2: Use",
  "3: Selection", "4: Contribution"
)[as.integer(evidence_counts$Level) + 1]
names(evidence_counts)[3] <- "Records"
knitr::kable(evidence_counts, caption = "Coded records for the selected fish.")
```

# Load the posterior draws

The repository includes the 180,000 retained draws from the FishKelp v7 fit.
Keeping the model fit separate makes this report quicker and easier to follow.
The full fitting command appears near the end of the document.

```{r posterior-draws}
posterior_draws_path <- file.path(
  output_dir, "strongest_fish_posterior_draws_v7.rds"
)
posterior_fit <- readRDS(posterior_draws_path)
posterior_draws <- posterior_fit$draws
```

# Posterior summaries

The summary below is calculated directly from the retained draws.

```{r posterior-summary}
summarise_route <- function(values, route) {
  data.frame(
    Species = selected_species,
    Route = route,
    Mean = mean(values),
    Median = median(values),
    `2.5%` = unname(quantile(values, 0.025)),
    `97.5%` = unname(quantile(values, 0.975)),
    check.names = FALSE
  )
}

posterior_summary <- rbind(
  summarise_route(posterior_draws$Trophic, "Trophic"),
  summarise_route(posterior_draws$Habitat, "Habitat")
)

summary_csv_path <- file.path(
  output_dir, paste0(species_slug, "_posterior_summary_v7.csv")
)
write.csv(posterior_summary, summary_csv_path, row.names = FALSE)

knitr::kable(
  posterior_summary,
  digits = 3,
  caption = "Posterior expected evidence reach in the West (kelp) context."
)
```

# Prior-predictive reference

Zoë's existing figures compare a blue dashed prior curve with a red posterior
curve. The same comparison is useful here, provided that "prior" has a precise
meaning. The code below samples the prior-predictive expected reach implied by
the actual FishKelp v7 priors. It includes the priors on route-specific
coefficients, the species effect and the ordered cut-points. It does not reuse a
posterior as a prior and does not simulate observations from a selected Beta
curve.

```{r prior-predictive-draws}
model_data <- readRDS(model_data_path)
species_index <- match(selected_species, model_data$species_lab)

set.seed(20260819)
n_prior_draws <- 50000L
nu0 <- 40
tau0 <- 0.7
cut_sd <- 5

# Under Sigma ~ inverse-Wishart(nu0, S0) and u | Sigma ~ N(0, Sigma),
# a single species effect has a multivariate t marginal distribution.
prior_df <- nu0 - model_data$D + 1
s0_diagonal <- tau0^2 * (nu0 - model_data$D - 1)
species_effect_scale <- sqrt(s0_diagonal / prior_df)

draw_prior_location <- function(route_index) {
  coefficient_draws <- matrix(
    rnorm(
      n_prior_draws * model_data$P,
      sd = rep(model_data$prior_scale, each = n_prior_draws)
    ),
    nrow = n_prior_draws,
    ncol = model_data$P
  )
  fixed_location <- drop(
    coefficient_draws %*% model_data$X_ref[species_index,]
  )
  species_effect <- species_effect_scale * rnorm(n_prior_draws) /
    sqrt(rchisq(n_prior_draws, df = prior_df) / prior_df)
  fixed_location + species_effect
}

draw_prior_cutpoints <- function() {
  free_cutpoints <- matrix(
    abs(rnorm(
      n_prior_draws * (model_data$C - 2L),
      sd = cut_sd
    )),
    nrow = n_prior_draws
  )
  free_cutpoints <- t(apply(free_cutpoints, 1, sort))
  cbind(0, free_cutpoints)
}

expected_reach <- function(location, cutpoints) {
  rowSums(1 - pnorm(cutpoints - location))
}

prior_draws <- data.frame(
  Trophic = expected_reach(
    draw_prior_location(1), draw_prior_cutpoints()
  ),
  Habitat = expected_reach(
    draw_prior_location(2), draw_prior_cutpoints()
  )
)
```

# Prior and posterior distributions

The next figure follows the presentation used in
`Posteriors_files/figure-html`: blue dashed curves are prior distributions and
red solid curves are posterior distributions. Unlike the earlier example
figures, both curves here are generated by the fitted FishKelp model and use the
evidence-reach scale.

```{r prior-posterior-plot}
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Prior-predictive and posterior distributions of trophic and habitat evidence reach for Pachymetopon blochii."
# Convert the wide draws into the long form expected by ggplot2.
long_draws <- function(draws, distribution) {
  long <- stack(draws[c("Trophic", "Habitat")])
  data.frame(
    Route = long$ind,
    Distribution = distribution,
    Reach = long$values
  )
}

comparison_long <- rbind(
  long_draws(prior_draws, "Prior"),
  long_draws(posterior_draws, "Posterior")
)
comparison_long$Route <- factor(
  comparison_long$Route,
  levels = c("Trophic", "Habitat")
)
comparison_long$Distribution <- factor(
  comparison_long$Distribution,
  levels = c("Prior", "Posterior")
)

level_labels <- c("None", "Co-occurrence", "Use", "Selection", "Contribution")
comparison_colours <- c(Prior = "#1F4AFF", Posterior = "#E31A1C")
comparison_linetypes <- c(Prior = "dashed", Posterior = "solid")

p_prior_posterior <- ggplot2::ggplot(
  comparison_long,
  ggplot2::aes(
    x = Reach,
    colour = Distribution,
    linetype = Distribution
  )
) +
  ggplot2::geom_density(linewidth = 0.9, adjust = 1.1) +
  ggplot2::facet_wrap(
    ~Route,
    nrow = 1,
    scales = "free_y",
    labeller = ggplot2::labeller(
      Route = c(
        Trophic = "Trophic evidence reach",
        Habitat = "Habitat evidence reach"
      )
    )
  ) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::scale_linetype_manual(values = comparison_linetypes) +
  ggplot2::scale_x_continuous(
    breaks = 0:4,
    labels = level_labels,
    limits = c(0, 4),
    expand = ggplot2::expansion(mult = c(0.01, 0.01))
  ) +
  ggplot2::labs(
    title = paste("Prior and posterior distributions for", selected_species),
    subtitle = "FishKelp v7 ordered-probit model",
    x = "Expected evidence level",
    y = "Density",
    colour = NULL,
    linetype = NULL
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    strip.background = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1),
    legend.position = "top",
    plot.title.position = "plot"
  )

prior_posterior_png <- file.path(
  output_dir, paste0(species_slug, "_prior_vs_posterior_v7.png")
)
prior_posterior_pdf <- file.path(
  output_dir, paste0(species_slug, "_prior_vs_posterior_v7.pdf")
)
ggplot2::ggsave(
  prior_posterior_png, p_prior_posterior,
  width = 10, height = 5, dpi = 300
)
ggplot2::ggsave(
  prior_posterior_pdf, p_prior_posterior,
  width = 10, height = 5
)

for (route_name in levels(comparison_long$Route)) {
  single_comparison <- ggplot2::ggplot(
    comparison_long[comparison_long$Route == route_name,],
    ggplot2::aes(
      x = Reach,
      colour = Distribution,
      linetype = Distribution
    )
  ) +
    ggplot2::geom_density(linewidth = 0.9, adjust = 1.1) +
    ggplot2::scale_colour_manual(values = comparison_colours) +
    ggplot2::scale_linetype_manual(values = comparison_linetypes) +
    ggplot2::scale_x_continuous(
      breaks = 0:4,
      labels = level_labels,
      limits = c(0, 4),
      expand = ggplot2::expansion(mult = c(0.01, 0.01))
    ) +
    ggplot2::labs(
      title = paste(route_name, "reach:", selected_species),
      subtitle = "Blue dashed: prior; red solid: posterior",
      x = "Expected evidence level",
      y = "Density",
      colour = NULL,
      linetype = NULL
    ) +
    ggplot2::theme_classic(base_size = 12) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 25, hjust = 1),
      legend.position = "top"
    )

  route_comparison_file <- file.path(
    output_dir,
    paste0(
      species_slug,
      "_",
      tolower(route_name),
      "_prior_vs_posterior_v7.png"
    )
  )
  ggplot2::ggsave(
    route_comparison_file, single_comparison,
    width = 7.5, height = 4.8, dpi = 300
  )
}

p_prior_posterior
```

# Interpretation

For *`r selected_species`*, the trophic posterior mean is
`r sprintf("%.3f", posterior_summary$Mean[posterior_summary$Route == "Trophic"])`
(95% credible interval
`r sprintf("%.3f--%.3f", posterior_summary[["2.5%"]][posterior_summary$Route == "Trophic"], posterior_summary[["97.5%"]][posterior_summary$Route == "Trophic"])`).
The habitat posterior mean is
`r sprintf("%.3f", posterior_summary$Mean[posterior_summary$Route == "Habitat"])`
(95% credible interval
`r sprintf("%.3f--%.3f", posterior_summary[["2.5%"]][posterior_summary$Route == "Habitat"], posterior_summary[["97.5%"]][posterior_summary$Route == "Habitat"])`).

These are relative evidence-reach estimates, not probabilities that the fish
is kelp-dependent. *P. blochii* is the strongest fish in the stated overall and
trophic rankings, but its posterior remains concentrated near the lower part of
the hierarchy. The result supports describing it as the leading fish candidate
for a kelp relationship in this evidence base. It does not support calling it kelp-
dependent: no fish record reaches Contribution or Dependence under the review
criteria.

# Reproduction and outputs

Render from the repository root with:

```{bash, eval=FALSE}
quarto render Bayesian_framework/FishKelp_strongest_fish_posteriors.qmd
```

Run the full model after a data change with:

```{bash, eval=FALSE}
Rscript R/run_fishkelp_v7_suite.R
```

The model script updates the retained focal draws. Render this report again to
update the figures. The HTML report is written beside this `.qmd`. The combined
PNG and PDF, route-specific PNGs, posterior draws and numerical summary are written to
`Bayesian_framework/posterior_outputs/`.
