---
title: "Kelp Forest Dependence of South African Fish Species"
author: "AJ Smit, Zoë-Angelique Petersen & David Dyer"
pdf-engine: lualatex 
format:
  html: 
    df_print: paged
    toc: true 
    toc_depth: 3
    theme: journal 
    highlight: zenburn 
    number_sections: true
    code_folding: hide 
    fig_width: 7 
    fig_height: 5
---

# What is this?

This qmd in particular will be assessed to assign informative prior to assess habitat dependency ranking of line-fish on kelp forests. Currently it consists of 7 questions (parameters) in totality:

* q1-3 forms the basis of kelp accessibility (these questions will place species in regions near or outside of kelp forests, without inferences on kelp occurrence or habitat use by species)
* q4-5 forms the basis of kelp occurrence - this places importance on species occurrence in kelp and the frequency their occurrence in kelp forests (leaning towards occurrence and selection of species in habitats)
* q6-7 forms the basis for use and dependency - the questions will draw on use of kelp forests, if there is a repeated, significant or exclusive use of kelp (allows conclusions of dependency to be drawn). Also includes life-stage history, as ontogenetic migrations can occur across habitats, places a greater level of level into the model.

The priors for this section will be set-up in two methods:
* Step 1: provides the parameters and values for Beta distributions for each parameter, which is posed as a question.

* Step 2: combines the individual parameters into a single integrated weighted prior parameter. (the weighting for each is discussed in that section and the equations are provided)

NB! This qmd will only comprise the prior set-up. The associated function code that will be used to assess species habitat priors will be in a separate qmd. This is for ease of later use and for code updates such that users can read the code and the framework alongside each other, instead of scrolling up and down to get the relevant information. 


# Habitat Dependency Ranking (HDR)

The Bayesian analysis framework aims to determine the HDR of line-fish on kelp forests. This will follow a methodological structure, of which priors will be first be set-up. The priors selected in this section will infer knowledge on the possible interactions line-fish have with kelp forests. Due to the lack of information there are with certain fish species, uncertainties will also be taken into consideration to provide a probabilistic estimate of the interaction occurring. Priors will be set up to account for this uncertainty and will later be integrated to include the different categories to display an overall HDR of line-fish on kelp forests.

## Priors

The priors are categories that will be used to determine the possible interaction between line-fish of kelp forests. These priors will encompass factors that will touch on the boundaries and extent of fish occurrence along the coast, along with the degree of habitat use of kelp forests by line-fish. These priors were selected based on the knowledge sourced from literature sources.

The prior represents our initial *beliefs* about a parameter's effect on the HDR estimate before observing the current data. In this context, the priors will encode our expert knowledge and assumptions about the TDR based on the answers to *Questions 1-7* in the Survey Questionnaire. These priors will then be updated with the data (likelihood) to produce the posterior distribution, which reflects our updated beliefs about the TDR.

Each question provides qualitative information that can be quantitatively translated into prior distributions for the HDR.

### Step 1: Assigning priors selection

The priors play a big role in shaping the posterior distributions. The type of prior---whether **Diffuse**, **Weakly Informative**, or **Informative**---reflects different levels of prior knowledge or assumptions about the parameter before seeing the data. Pay careful attention to assigning priors. A diffuse or non-informative prior is chosen to exert minimal influence on the posterior distribution and allows the data to drive the posterior in the absence of strong prior beliefs. Weakly informative priors provide some guidance without strongly biasing the posterior. They reflect a minimal level of knowledge or reasonable assumptions about the parameter's likely range without assuming high confidence in specific values. Informative priors are used when there is strong prior knowledge or beliefs about the parameter.


#### **Question 1: Geographic Distribution**

Because the distribution of kelp forests and the availability of kelp as a food source vary between the west and east coasts (*Ecklonia maxima* at De Hoop) regions, fish species that are distributed in different regions along the South African coast may have different HDRs.

Select the option that best describes the geographical distribution of the fish species along the South African coast in relation to the distribution of kelp forests (*Ecklonia maxima* and/or *Laminaria pallida*). This information will help assess the species' potential HDR on kelp.

***Choices: \[West \| East \| Both\]***

-   **\[West\]**: The species is *exclusively* found in the western region of the South African (and/or Namibian) coast, which is dominated by kelp forests (*Ecklonia maxima* and/or *Laminaria pallida*). Therefore, it is *likely* that the species has a *higher HDR* on kelp, potentially even indicating an *obligate dependency* on kelp-derived carbon.
-   **\[East\]**: The species is *exclusively* found along the coastline *east of the eastern limit* of kelp forests (east of De Hoop), where *Ecklonia maxima* does not occur, and kelp forests are absent (but there's still *Ecklonia radiata*). Consequently, the species is expected to have a HDR *close to zero* (not zero due to chance of accessing *E. radiata*), indicating *no to minimal dependency* on kelp-derived carbon.
-   **\[Both\]**: The species is found in *both* the western region (with kelp forests) and the eastern region (without kelp forests). As a result, the species may have an *intermediate HDR*, suggesting a *facultative dependency* on kelp-derived carbon, where the species utilises kelp when available but can also thrive without it.

Translation into Priors:

To model a parameter like HDR, which is a proportion bounded between 0 and 1 (or 0% to 100%), use a Beta distribution. It has a flexible shape and so will allow for various levels of certainty and mean values. The Beta distribution has two shape parameters, $\alpha$ and $\beta$, which determine the mean (@eq-equation1) and variance (@eq-equation2) of the distribution:

$$ \mu = \frac{\alpha}{\alpha + \beta} $$ {#eq-equation1}

$$ \sigma^2 = \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)} $$ {#eq-equation2} 

Example profiles of the Beta distribution for different values of $\alpha$ and $\beta$ are shown in @fig-fig1:

```{r beta_distribution}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Parameters of the Beta distribution."
#| label: fig-fig1
#| fig.align: center
#| fig-height: 9.5
#| fig-width: 12
#| fig-pos: "t"
#| fig.asp: 0.85

# Load necessary libraries
library(tidyverse)
library(ggdist)

# Define a function to calculate and plot the Beta distribution for given alpha and beta values
plot_beta_distribution <- function(alpha, beta) {
  # Create a sequence of values for the TDR (0 to 1)
  hdr_values <- seq(0, 1, length.out = 1000)
  
  # Calculate the Beta distribution density
  density_values <- dbeta(hdr_values, alpha, beta)
  
  # Calculate mean and variance
  mean_value <- alpha / (alpha + beta)
  variance_value <- (alpha * beta) / ((alpha + beta)^2 * (alpha + beta + 1))
  
  # Create a data frame for plotting
  plot_data <- data.frame(HDR = hdr_values, Density = density_values)
  
  # Create the plot
  plot <- ggplot(plot_data, aes(x = HDR, y = Density)) +
    geom_line(color = "blue", size = 1) +
    labs(title = paste("Beta Distribution (α =", alpha, ", β =", beta, ")"),
         subtitle = paste("Mean =", round(mean_value, 4), "| Variance =", round(variance_value, 4)),
         x = "Habitat Dependency Rank (HDR)",
         y = "Density") +
    theme_minimal()
  
  return(plot)
}

# Define a list of parameter combinations to explore
params <- list(
  list(alpha = 1, beta = 1),
  list(alpha = 2, beta = 5),
  list(alpha = 4, beta = 1),
  list(alpha = 1, beta = 19),
  list(alpha = 2, beta = 2),
  list(alpha = 5, beta = 5),
  list(alpha = 10, beta = 2),
  list(alpha = 5, beta = 10)
)

# Generate and display plots for each parameter combination
plots <- lapply(params, function(p) plot_beta_distribution(p$alpha, p$beta))

# Arrange the plots in a grid
library(gridExtra)
do.call(grid.arrange, c(plots, ncol = 2))
```

We can solve for α and β given the desired mean and variance.

For example, for the 'West' region, our desired mean is $\mu = 0.8$, which reflects a higher HDR. We choose a variance that indicates more uncertainty (because a species simply being confined to this region offers little certainty that it has a HDR reflective of kelp usage). Since we want a wider distribution (higher uncertainty), we can choose $\alpha$ and $\beta$ to be smaller. A Beta(4,1) distribution can be chosen to reflect a high HDR (mean = 0.8 from @eq-equation3) with moderate uncertainty (variance \~0.0267 from @eq-equation4). The calculations are as follows:

$$\mu = \frac{4}{4 + 1} = 0.8$$ {#eq-equation3}

$$\sigma^2 = \frac{4 \times 1}{(4 + 1)^2 (4 + 1 + 1)} = \frac{4}{25 \times 6} = \frac{4}{150} \approx 0.0267$$ {#eq-equation4}

The priors for the 'West,' 'East,' and 'Both' geographic regions are therefore:

| **Region** | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|------------|----------------------------|--------------|-------------------|
| **West**   | α = 4, β = 1               | μ = 0.8      | σ² ≈ 0.0267       |
| **East**   | α = 1, β = 19              | μ = 0.05     | σ² ≈ 0.00226      |
| **Both**   | α = 2, β = 2               | μ = 0.5      | σ² = 0.05         |

Visualising these priors helps us understand how they reflect our prior beliefs about the HDR. The priors are shown in @fig-fig2 and @fig-fig3:

```{r question1-priors}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Plot of prior distributions for the geographic distribution effect of fish regional distribution on HDR."
#| label: fig-fig2
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the Beta distributions for the priors
priors <- data.frame(
  x = seq(0, 1, length.out = 1000),
  west = dbeta(seq(0, 1, length.out = 1000), 4, 1),
  east = dbeta(seq(0, 1, length.out = 1000), 1, 19),
  both = dbeta(seq(0, 1, length.out = 1000), 2, 2)
)

# Plot the prior distributions
ggplot(priors, aes(x = x)) +
  geom_line(aes(y = west, color = "West"), size = 1) +
  geom_line(aes(y = east, color = "East"), size = 1) +
  geom_line(aes(y = both, color = "Both"), size = 1) +
  scale_color_manual(values = c("West" = "blue", "East" = "red", "Both" = "green")) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Geographic Distribution Effect",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.8))

```


```{r question1-priorsdis}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Plot of prior distributions for the geographic distribution effect of fish regional distribution on HDR."
#| label: fig-fig3
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  West = list(alpha = 4, beta = 1),
  East = list(alpha = 1, beta = 19),
  Both = list(alpha = 2, beta = 2)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("West", "East", "Both")))

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Depth Effect",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8))
```

#### **Question 2: Depth**

The priors for the depth effect on HDR will have to reflect the fact that species that occur at the same depth as kelp forests are more likely to have a higher HDR on kelp, and as we go into deeper water, the HDR on kelp is expected to decrease. Literature often reports depth as a range, broad category or maximum depth rather than consistent point estimates (depending on the assessment done). Therefore depth will be classified into categories rather than exact depth values. The depth itself does not represent occurrence in kelp, but reflects the potential for species to overlap vertically with kelp forests.

***Choices: \[Unknown \| Shallow nearshore \| Deeper nearshore \| Mid-water \| Deep water \]***

- **\[Unknown\]**: *No reliable or reported depth* parameters for species occurrences in the water column. There will be an expected *low HDR*, as unknowns have no information to suggest an overlap in kelp boundaries
- **\[Shallow nearshore\]**: Species are reported in *intertidal, surf-zone, shallow subtidal zones* with depth ranges of *0-10m* in the water column. There will be an expected *moderately high HDR* as there is a stronger opportunity for a possible overlap with canopy kelp systems.
- **\[Deeper nearshore\]**: Species are reported in depths from approximately *10-30m* in the water column. There is an expected *moderately high HDR*, as there is an opportunity for possible overlap with kelp forest systems.
- **\[Mid-water\]**: Species are reported mainly in offshore reefs, shelf waters or mid-water habitats that range in depth from *30-60m*. There is an expected *moderately low HDR*, as there is a limited potential of overlap, with possible deeper kelp forest systems which are not common. and episodic movements between vertical columns.   
- **\[Deep water\]**: Species has a reported occurrence in deeper waters from which are greater than 60m. There is an expected *very low HDR* as there is a very low potential for kelp overlap.
- **\[Many\]**: Species are reported at greater depth ranges encompassing depths from two or more options. There is an expected *moderately low HDR* as species occur outside of ~E. maxima~ distribution boundaries, while occurring in depths suitable for ~E. radiata~  

NB! In kelp boundaries use the shallowest documented ecological relevant depth rather than the maximum depth. If a species occurs in shallow waters but also deeper waters but along the 'West' coast i.e. 5-150m, then report shallow option. For species outside of kelp boundaries use 'Many'.

*Step 1: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* Unknown ~ 0.5 (low HDR)
* Shallow nearshore ~ 0.7 (very high HDR)
* Deep nearshore ~ 0.6 (high HDR)
* Mid-water ~ 0.35 (moderately low HDR)
* Deep water ~  0.125 (low HDR)
* Many ~ 0.3 (moderate HDR)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. Unknown (target mean = 0.5)
Using mean = α/(α+β) = 0.5


$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5 α + 0.5 β
$$\

$$
0.5α = 0.5 β
$$\

$$
α = β
$$\

I selected β = 1 and giving α = 1\ (uninformative prior)

2. Shallow nearshore (target mean = 0.7):
Using mean = α/(α+β) = 0.7


$$
\frac{α}{(α+β)} = 0.7
$$\

$$
α = 0.7 \cdot (α+β)
$$\

$$
α = 0.7α + 0.7β
$$\

$$
0.3α = 0.7β
$$\

$$
α = \frac{7}{3}β
$$\

I selected β = 3 and α = 7\

3. Deeper nearshore (target mean = 0.6):

* Using similar equations: α/(α+β) = 0.6

$$
\frac{α}{(α+β)} = 0.6
$$\

$$
α = 0.6 \cdot (α+β)
$$\

$$
α = 0.6α + 0.6β
$$\

$$
0.4α = 0.6β
$$\

$$
α = 1.5β
$$\

I selected β = 4 and α = 6

4. Mid-water (target mean = 0.3):

* Using mean = α/(α+β) = 0.3

$$
\frac{α}{(α+β)} = 0.3
$$\

$$
α = 0.3 \cdot (α+β)
$$\

$$
α = 0.3α + 0.3β
$$\

$$
0.7α = 0.3β
$$\

$$
α = 0.429β
$$\

I selected β = 7 and α = 3

5. Deep water (target mean = 0.125):

* Using mean = α/(α+β) = 0.125

$$
\frac{α}{(α+β)} = 0.125
$$\

$$
α = 0.125 \cdot (α+β)
$$\

$$
α = 0.125α + 0.125β
$$\

$$
0.875α = 0.125β
$$\

$$
α = 0.143β
$$\

I selected β = 7 and α = 1

6. Many (target mean = 0.35):
This represents a very strong prior belief that the dependency score is close to one

* Using mean = α/(α+β) = 0.35

$$
\frac{α}{(α+β)} = 0.35
$$\

$$
α = 0.35 \cdot (α+β)
$$\

$$
α = 0.35α + 0.35β
$$\

$$
0.65α = 0.35β
$$\

$$
α = 0.538β
$$\

I selected β = 6 and α = 3.228


Translated into Priors:

| **Depth**               | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|-------------------------|----------------------------|--------------|-------------------|
| **Unknown**             | α = 1, β = 7               | μ = 0.5      | σ² ≈ 0.0833       |
| **Shallow nearshore**   | α = 7, β = 3               | μ = 0.7      | σ² ≈ 0.0191       |
| **Deeper nearshore**    | α = 6, β = 4               | μ = 0.6      | σ² ≈ 0.0218       |
| **Mid-water**           | α = 3, β = 7               | μ = 0.3      | σ² ≈ 0.0191       |
| **Deep water**          | α = 1, β = 7               | μ = 0.125    | σ² ≈ 0.0122       |
| **Many**                | α = 3.228, β = 6           | μ = 0.35     | σ² ≈ 0.0222       |

A visualisation of these priors is given in @fig-fig4:

```{r question2-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of species depth profiles along the South African coast."
#| label: fig-fig4
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  #Unknown = list(alpha = 1, beta = 1), (uninformative prior)
  Shallow_nearshore = list(alpha = 7, beta = 3),
  Deeper_nearshore = list(alpha = 6, beta = 4),
  Mid_water = list(alpha = 3, beta = 7),
  Deeper_water = list(alpha = 1, beta = 7),
  Many = list(alpha = 3.228, beta = 6)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Kelp-use",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```

#### **Question 3: Coastal position**

The priors for the distance from the shore effect on HDR will have to reflect the fact that species that occur closer to the shore are more likely to have a higher HDR on kelp, and as we move further offshore, the HDR on kelp is expected to decrease. The priors for this question are as follows:

***Choices: \[ Unknown \|Immediate nearshore \| Nearshore-offshore continuum \| Offshore \| Pelagic \]*** 

- **\[Unknown\]**: There is *no clear coastal position* described for species in South African waters. There is an expected *low HDR* as no inferences can be drawn about a species accessibility to kelp forests.
- **\[Immediare nearshore\]**: Species occupies zones *closer to the shore* such as intertidal, surf-zone, tidal pool, estuary mouth, inshore, nearshore, shallow coastal reef or rocky shore. There will be an expected *moderately high HDR* as these zones offer the highest opportunity for species accessibility to kelp forests.
- **\[Offshore\]**: The species occurs in habitats *further from the shore* which include zones noted as offshore reefs, shelf waters, outer coastal zones, continental shelf, deeper coastal waters. There is an expected *low HDR* as there is a limited opportunity for species to access kelp forest systems.
- **\[Nearshore-offshore\]**: Species occupies zones in between nearshore and offshore regions, together with both zones. These zones include habitats such as inshore and offshore reefs, coastal waters, bays and embayments, reef-associated migrations can be included in this section. There is an expected *moderate HDR* as there is a possible opportunity for species to access kelp forest environments.  
- **\[Pelagic\]**: Species occupies zones that are *off the continental shelf*. These include zones such as pelagic, epipelagic, oceanic, open water column and offshore water. There is an expected *low HDR* as species has the lowest opportunity to access kelp forests.
- **\[Many\]**: Species occupies nearshore and/or offshore zones together with zones that are *off the continental shelf*. There is an expected *moderately low HDR* as species has a low opportunity to access kelp forests.

Translated into Priors:

| **Coastal position**      | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|---------------------------|----------------------------|--------------|-------------------|
| **Unknown**               | α = 1, β = 1               | μ ≈ 0.5      | σ² ≈ 0.0833       |
| **Immediate nearshore**   | α = 7, β = 3               | μ = 0.7      | σ² ≈ 0.0191       |
| **Offshore**              | α = 3, β = 7               | μ ≈ 0.3      | σ² ≈ 0.0191       |
| **Nearshore-offshore**    | α = 5, β = 5               | μ ≈ 0.5      | σ² ≈ 0.0227       |
| **Pelagic**               | α = 1, β = 7               | μ ≈ 0.125    | σ² ≈ 0.0122       |
| **Many**                | α = 3.228, β = 6           | μ = 0.35     | σ² ≈ 0.0222       |

A visualisation of these priors is given in @fig-fig5:

```{r question3-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of species coastal position along the South African coast."
#| label: fig-fig5
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  #Unknown = list(alpha = 1, beta = 1), (uninformative prior)
  Immediate_nearshore = list(alpha = 7, beta = 3),
  Offshore = list(alpha = 3, beta = 7),
  Nearshore_offshore = list(alpha = 5, beta = 5),
  Pelagic = list(alpha = 1, beta = 7),
  Many = list(alpha = 3.228, beta = 6)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Kelp-use",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```


#### **Question 4: Habitat occurrence**

In this question you are asked to select the answer that best describes the habitat of the fish species in relation to kelp forests. Answering 'In', 'Adjacent', 'Near', or 'Both' pertains to instances when you answered 'West' in Question 1. Answering 'Away' applies only to when you answered 'East' in Question 1.

***Choices: \[In \| Adjacent \| Both \| Away\]***

-   **\[In\]**: The species is *found within* kelp forests, where it is likely to have a *higher Trophic Dependency Rank (HDR)* on kelp.
-   **\[Adjacent\]**: The species is *found in habitats adjacent to* kelp forests, where it may have a *moderate HDR* on kelp.
- **\[Near\]**: The species is *found in regions near to* kelp forests, where it may have a *moderate HDR* on kelp.  
-   **\[Both\]**: The species is found in *both* kelp forests and adjacent areas, suggesting an *intermediate HDR* on kelp-derived carbon.
-   **\[Away\]**: The species is *found away from* kelp forests, where it is expected to have a *low HDR* on kelp, indicating *no dependency* on kelp-derived carbon.

Translation into Priors:

| **Habitat**  | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|--------------|----------------------------|--------------|-------------------|
| **In**       | α = 8, β = 1               | μ ≈ 0.8889   | σ² ≈ 0.00988      |
| **Adjacent** | α = 4, β = 1               | μ = 0.8      | σ² ≈ 0.02667      |
| **Both**     | α = 6, β = 1               | μ ≈ 0.8571   | σ² ≈ 0.01531      |
| **Away**     | α = 5, β = 85              | μ ≈ 0.0556   | σ² ≈ 0.0005765    |

A visualisation of these priors is given in @fig-fig6:

```{r question4-priors}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distributions for the habitat occurrence in kelp forests."
#| label: fig-fig6
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  In = list(alpha = 8, beta = 1),
  Adjacent = list(alpha = 2, beta = 2),
  Both = list(alpha = 7.7, beta = 3.3),
  Away = list(alpha = 1, beta = 19)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Habitat Effect",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8))
```

*\[Note: We would have to run a sensitivity analysis to determine the effect of the priors. For example, in a situation where a species in Question 1 is scored as 'West' and in Question 2 as 'In', the priors for the geographic distribution effect and the habitat effect would be combined to determine the overall effect on HDR. Since the species is found in the west and within kelp forests, the priors for the geographic distribution effect and the habitat effect would be multiplied to determine the overall effect on HDR, which could be close to 1 making it obligative (the fish only occurs directly within kelp forests, but at this point we still have no insight about whether it is a trophic or habitat dependence, or both). This process would be repeated for all possible combinations of answers to Questions 1 and 2.\]*

#### **Question 5: Frequency of occurrences**

In this question you are asked to select the answer that best describes the use of kelp forests by species. Answering either 'No use', 'Incidental use', 'Regular use', 'Strong use' or 'Near-exclusive use' pertains to instances when you answered 'West' in Question 1 and 'In' in Question 4. Answering 'No use' applies to all categories. 

***Choices: \[No occurrence \| Incidental occurrence \| Regular occurrence \| Significant occurrence \| Near-exclusive occurrence \]***

- **\[No occurrence\]**: The species has *no documented or reported occurrence* of a species in kelp forests . There is an expected *low HDR*, whereby there is *no kelp-link* for habitat provisioning.
- **\[Incidental occurrence\]**: The species has *rare or isolated reports* of kelp occurrence. There is an expected *low HDR*, whereby there is a limited occurrence of species in kelp forests.
- **\[Regular occurrence\]**: The species has *repeated reports of occurrence* of a species in kelp forests across many sites, seasons or individuals. There is an expected *moderate HDR* indicating a possible substantial use of kelp for habitat provisioning.
- **\[Significant occurrence\]**: The species has a *disproportionate occurrence* of species in kelp forests relative to alternative ecosystems for habitat use. There is an expected *relatively high HDR*, indicating a possible stronger use of kelp for improved species fitness.
- **\[Near-exclusive occurrence\]**: The species has *occurrence reports mainly in kelp forests* across several sample sites and years, with alternative ecosystems rarely used. There is an expected *high HDR*, whereby there is an almost exclusive use of kelp for a possible improved species fitness through habitat provisioning. 

Translated into Priors:

| **Frequency of occurrence**   | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|-------------------------------|----------------------------|--------------|-------------------|
| **No occurrence**             | α = 1, β = 6               | μ ≈ 0.143    | σ² ≈ 0.0153       |
| **Incidental occurrence**     | α = 3, β = 7               | μ = 0.3      | σ² ≈ 0.0191       |
| **Regular occurrence**        | α = 6, β = 5               | μ ≈ 0.55     | σ² ≈ 0.0207       |
| **Strong occurrence**         | α = 7, β = 2               | μ ≈ 0.857    | σ² ≈ 0.0173       |
| **Near-exclusive occurrence** | α = 9, β = 1               | μ ≈ 0.9      | σ² ≈ 0.0082       |

A visualisation of these priors is given in @fig-fig7:

```{r question5-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distribution of species frequency of occurrence in kelp forests"
#| label: fig-fig7
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  No_occurrence = list(alpha = 1, beta = 6),
  Incidental_occurrence = list(alpha = 3, beta = 7),
  Regular_occurrence = list(alpha = 6, beta = 5),
  Significant_occurence = list(alpha = 7, beta = 2),
  Near_exclusive_occurrence = list(alpha = 9, beta = 1)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Kelp-use",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8))

```

#### **Question 6: Life-stage use**

In this question you are asked to select the answer that best describe the species life-stage documented in kelp forests. Answering 'Juveniles', 'Adults', 'Both', 'KR juvenile', 'KR adult' or 'KR both'  applies to instances when you answered 'West' Question 1 and 'In' or 'Both' in Question 4. 'KR' indicates a kelp-restricted occurrence or use in species, whereas the other options where 'KR' is not applied occurs in several habitats including kelp forests. 'None' documented applies to all categories. 'Unsure' applies to instances where species were sampled within kelp boundaries, but you unsure of whether the individuals were adults or juveniles.

***Choices: \[None \| Unsure \| Juvenile \| Adult \| Both \| KR juvenile\| KR adult\| KR both \]***

- **\[None\]**: The species has no life-stage that occurs or uses kelp forests for habitat provisioning. There is an expected *low HDR* as species there is no evidence or reports of kelp use.
- **\[Unsure\]**: The species has a documented occurrence or use in kelp forests, but *no life-stage was reported*. The species has an expected *moderate HDR* in kelp forest as there is substantial evidence or reports of kelp forest use, but not detailed reports.
- **\[Juvenile\]**: The species occurs or uses kelp *only during their juvenile life-stage*. There is an expected *moderately high HDR* as there is evidence to support kelp use, but *not enough evidence of a restriction to kelp*. 
- **\[Adult\]**: The species occurs or uses kelp *only during their adult life-stage*. There is an expected *moderately high HDR* as there is evidence to support kelp use, but *not enough evidence of a restriction to kelp*.
- **\[Both\]**: The species occurs or uses kelp *across all life-stages*. There is an expected *moderately high HDR* as there is evidence to support kelp use, but *not enough evidence of a restriction to kelp*.
- **\[KR juvenile\]**: The species only occurs or uses kelp *during their juvenile life-stage*. There is an expected *high HDR* as there is evidence to support kelp use.
- **\[KR adult\]**: The species only occurs or uses kelp *during their adult life-stage*. There is an expected *high HDR* as there is evidence to support kelp use.
- **\[KR both\]**: The species only occurs or uses kelp *across all life-stages*. There is an expected *high HDR* as there is evidence to support kelp use.

Translated into Priors:

| **Habitat function**           | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|--------------------------------|----------------------------|--------------|-------------------|
| **None**                       | α = 1, β = 6               | μ ≈ 0.143    | σ² ≈ 0.0153       |
| **Unsure**                     | α = 6, β = 3               | μ ≈ 0.67     | σ² ≈ 0.0222       |
| **Juvenile**                   | α = 6, β = 3               | μ ≈ 0.67     | σ² ≈ 0.0222       |
| **Adult**                      | α = 6, β = 3               | μ ≈ 0.67     | σ² ≈ 0.0222       |
| **Both**                       | α = 6, β = 3               | μ ≈ 0.67     | σ² ≈ 0.0222       |
| **KR juvenile**                | α = 9, β = 1               | μ ≈ 0.9      | σ² ≈ 0.0082       |
| **KR adult**                   | α = 9, β = 1               | μ ≈ 0.9      | σ² ≈ 0.0082       |
| **KR both**                    | α = 9, β = 1               | μ ≈ 0.9      | σ² ≈ 0.0082       |

A visualisation of these priors is given in @fig-fig8:

```{r question6-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of reported habitat function by species across multiple life-stages"
#| #| fig.align: center
#| label: fig-fig8
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  None = list(alpha = 1, beta = 9),
  Unsure = list(alpha = 5, beta = 5),
  Juvenile = list(alpha = 5, beta = 4),
  Adult = list(alpha = 5, beta = 4),
  Both = list(alpha = 5, beta = 4),
  KR_juvenile = list(alpha = 10, beta = 1),
  KR_adult = list(alpha = 10, beta = 1),
  KR_both = list(alpha = 10, beta = 1)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Kelp-use",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```


#### **Question 7: Habitat-function**

In this question you are asked to select the answer that best described the habitat function of kelp forests for species. Answering 'Plausible mechanism', 'Habitat evidence', 'Comparative evidence', and 'Fitness evidence' pertains to instances when you answered 'West' in Question 1 and 'In' or 'Both' in Question 4. Answering 'None documented' applies to all categories. 

***Choices: \[None \| Plausible mechanism \| Reported evidence \| Comparative evidence \| Fitness evidence \| Dependence evidence \]***

- **\[None\]**: The species is *no documented*  habitat use of kelp forests for habitat provisioning. There is an expected *low HDR*, whereby there is *no kelp-link* for habitat provisioning.
- **\[Plausible mechanism\]**: The species is *no documented habitat use* of kelp forests but researchers suggest a *possible suitability* of habitats for species due to their occurrence in kelp and the forests meeting their requirements. Authors describe kelp forests as likely refuge or foraging areas. There is an expected *low HDR*, whereby there is a speculated use of kelp forests for habitat provisioning.
- **\[Reported evidence\]**: The species has *repeated reports of habitat use* of kelp forests for shelter, breeding, nursery and/or foraging grounds across many sites, seasons or individuals. There is a documented habitat use of kelp forests by species. There is an expected *moderately high HDR* indicating a habitat use of kelp for habitat provisioning.
- **\[Comparative evidence\]**: The species has a *disproportionate habitat use of kelp* for shelter, breeding, nursery and/ foraging grounds relative to alternative ecosystems for habitat use. There is an expected *relatively high HDR*, indicating a stronger use of kelp for improved species fitness.
- **\[Contribution evidence\]**: The species has a *documented ecological fitness linked to kelp forests* across several sample sites and years. There is an expected *high HDR*, whereby there is a contribution relationship present through habitat provisioning.
- **\[Dependence evidence\]**: The species has a *documented ecological fitness linked to kelp forests* across several sample sites and years. Kelp is linked to a species survival, growth, recruitment, condition and reproduction with a demonstrated demographic response in kelp loss, whereby alternative . There is an expected *high HDR*, whereby there is an almost exclusive use of kelp for improved species fitness through habitat provisioning.

*Step 1: Understanding the Ecological Relationship*\
The expected TDR of line fish will be high for individuals that have a greater kelp trophic function and smaller for line fish that have no kelp trophic function. 

- **\[None\]**: No or currently tested kelp or kelp-associated prey link. This places the species trophic pathway in alternative ecosystems. 
- **\[Plausible mechanism\]**: No documented habitat use, but researcher suggest a *possible suitability* of habitats for species occurrence and use through shelter and foraging. While there is a possibility, this is not supported by evidence, which places this choice as an uncertainty and future research gap at a low HDR.
- **\[Reported evidence\]**: The species has a repeated use of habitats for shelter, foraging, breeding or nursery areas, which form part one component of broad habitat use. The habitat functioning is moderate, as kelp forms only a part of their function but no ecological improvement in species fitness is noted.
- **\[Comparative evidence\]**: Species has a disproportionate kelp habitat use relative to alternative ecosystems. It indicates a selection of kelp forests.
- **\[Contribution evidence\]**: There is strong evidence and source justification for an ecological improvement in line-fish ecological fitness. However, species can undergo this contribution in alternative ecosystems if kelp forests were reduced.  
- **\[Dependence evidence\]**: There is strong evidence and source justification for an ecological improvement in line-fish ecological fitness. Without kelp forests present, there would be a decline in species abundance, whereby alternative ecosystems cannot replace the trophic pathway. 

*Step 2: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* None ~ 0.12 (low dependency)
* Common ~ 0.54 (moderate dependency)
* Significance ~ 0.75 (high dependency)
* Dependent ~ 0.9 (very high dependency)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. None (target mean = 0.102)
Using mean = α/(α+β) = 0.102


$$
\frac{α}{(α+β)} = 0.102
$$\

$$
α = 0.102 \cdot (α+β)
$$\

$$
α = 0.102 α + 0.102 β
$$\

$$
0.898α = 0.102 β
$$\

$$
α = 0.114β
$$\

I selected β = 8.8 for narrow variance, giving α $\approx$ 1\

2. Plausible evidence (target mean = 0.3):
Using mean = α/(α+β) = 0.3


$$
\frac{α}{(α+β)} = 0.3
$$\

$$
α = 0.3 \cdot (α+β)
$$\

$$
α = 0.3α + 0.3β
$$\

$$
0.7α = 0.3β
$$\

$$
α = 0.429β
$$\

I selected β = 7 and α = 3\

3. Reported evidence (target mean = 0.5):

* Using similar equations: α/(α+β) = 0.5

$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5α + 0.5β
$$\

$$
0.5α = 0.5β
$$\

$$
α = β
$$\

I selected β = 5 and α = 5

4. Comparative evidence (target mean = 0.636):

* Using mean = α/(α+β) = 0.636

$$
\frac{α}{(α+β)} = 0.636
$$\

$$
α = 0.636 \cdot (α+β)
$$\

$$
α = 0.636α + 0.636β
$$\

$$
0.364α = 0.636β
$$\

$$
α = 1.75β
$$\

I selected β = 4 and α = 7

5. Contribution evidence (target mean = 0.8):

* Using mean = α/(α+β) = 0.8

$$
\frac{α}{(α+β)} = 0.8
$$\

$$
α = 0.8 \cdot (α+β)
$$\

$$
α = 0.8α + 0.8β
$$\

$$
0.2α = 0.8β
$$\

$$
α = 4β
$$\

I selected β = 2 and α = 8

6. Dependence (target mean = 0.99):
This represents a very strong prior belief that the dependency score is close to one

* Using mean = α/(α+β) = 0.99

$$
\frac{α}{(α+β)} = 0.99
$$\

$$
α = 0.99 \cdot (α+β)
$$\

$$
α = 0.99α + 0.99β
$$\

$$
0.01α = 0.99β
$$\

$$
α = 99β
$$\

I selected β = 1 and α = 99


Translated into Priors:

| **Habitat function**           | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|--------------------------------|----------------------------|--------------|-------------------|
| **None**                       | α = 1, β = 8.8             | μ ≈ 0.102    | σ² ≈ 0.0085       |
| **Plausible mechanism**        | α = 3, β = 7               | μ = 0.3      | σ² ≈ 0.0191       |
| **Reported evidence**          | α = 5, β = 5               | μ ≈ 0.5      | σ² ≈ 0.0227       |
| **Comparative evidence**       | α = 7, β = 4               | μ ≈ 0.636    | σ² ≈ 0.0193       |
| **Contribution evidence**      | α = 8, β = 2               | μ ≈ 0.8      | σ² ≈ 0.0145       |
| **Dependence evidence**        | α = 99, β = 1              | μ ≈ 0.99     | σ² ≈ 0.0001       |

A visualisation of these priors is given in @fig-fig9:

```{r question7-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distribution of reported habitat function of species in kelp forests"
#| label: fig-fig9
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  None = list(alpha = 1, beta = 8.8),
  Plausible_mechanism = list(alpha = 3, beta = 7),
  Reported_evidence = list(alpha = 5, beta = 5),
  Comparative_evidence = list(alpha = 7, beta = 4),
  Contribution_evidence = list(alpha = 8, beta = 2),
  Dependence_evidence = list(alpha = 9, beta = 1)
)

# Create a sequence of HDR values from 0 to 1
hdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = hdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(hdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -HDR)

# Plot the densities
ggplot(density_long, aes(x = HDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for HDR",
       subtitle = "Kelp-use",
       x = "HDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```

### Step 2: Integrating the prior parameters

Since priors play an important role in shaping posteriors, I wanted to ensure prior Beta distributions for each parameter was carefully generated. These parameters are not single entities, but rather form collective insights of how line-fish use kelp forests. The first three questions (Q1-3) form part of the baseline assessment that places a species in kelp forest boundaries. Questions 4-6 provides insight into the use of kelp forests by line-fish, whether the occurrence is a mere overlap in geographical boundaries or if a use was reported across different life-stages. Question 7 provides greater clarity on kelp forest use: whether there would be a species population decline in response to kelp loss. Thus, each of these parameters hold different weights in the final outcome:\

#### Weights

As mentioned, each section will weigh differently in the final integration. The weights will add up to 1. 

- Since q1-3 forms the baseline of the assessment, placing a species in kelp boundaries, the weights of these will be low as they do not inform strong relationship but rather a possibility. They will be weighted low.
- Q4 places a species in kelp forests, it describes their occurrence in kelp forests, whether their occurrence was significant or near-exclusive relative to alternative habitats. This parameter will have a greater weight than q1-3 but not as high as it does not inform habitat use, rather species occurrence.
- Q5 places infers frequencies of species occurrence in kelp forest, whether observations are incidental, common, significant or near exclusive relative to alternative habitats. As it goes a bit deeper into species occurrences in kelp forest it will weigh more than q1-4.
- Q6 provides information on species life-stage use (i saw in the HDI they used species life-stage, i thought okay that's informative). Again this section refers to different life-stages that use or occur in kelp forests with options for all habitats and options for kelp restricted life stages. It will have a greater weight than q1-4, while maintaining the same weight at q5.
- Q7 will provide insights into habitat functioning by species, indicating whether there is a possibility that species can use habitats, if they do use kelp forests, if their use of kelp is significant, or whether they are dependent on kelp forests to under a continuation their life-cycle. This will have the greatest weight as it places the greatest insight into habitat use by line-fish.

#### How to read the Beta distributions?

* values of 1 indicating dependency on kelp forests by line-fish\
* values above 0.5 indicating a strong use of kelp\
* values below 0.5 indicating a weak kelp use\
* and 0 indicating no relationship to kelp forests. 

Thus, values closer to 1 on the beta distribution indicates a stronger relationship between line-fish and kelp.

In order to assess this, a weighted logarithmic pool of seven beta distributions will be used:

Weight log pool:
$$ \pi_{HDR,i} (\theta) \propto \prod_{q=1}^{7} \left[ \pi_q(\theta \mid \alpha_{qi}, \beta_{qi}) \right] w_q $$


where w~q~ $\ge$ 0 and:

$$ \sum_{q=1}^{7} w_q = 1 $$

For Beta distributions, this remains a beta distribution:

$$ \theta_{H,i}^{prior} \sim Beta(\alpha_i, \beta_i) $$

with:

$$ \alpha_i = 1 +\sum_{q=1}^{7} w_q(\alpha_{qi} - 1) $$

$$ \beta_i = 1 +\sum_{q=1}^{7} w_q(\beta_{qi} - 1) $$

because the weights sum to 1, this is equivalently:

$$ \alpha_i = \sum_{q=1}^{7} w_q \alpha_{qi} $$

$$ \beta_i = \sum_{q=1}^{7} w_q \beta_{qi} $$

Altogether, this generates a single integrated prior, whereby weighted pooling combines distributions using weighted mean and retains a Beta form.

Possible weights for each question:

| *Question*       | *Parameter*              | *Weight*        | *Role*                                        |
|------------------|--------------------------|-----------------|-----------------------------------------------|
| 1                | distribution range       | 0.05            | species accessibility to kelp forests         |
| 2                | depth profile            | 0.05            | species accessibility to kelp forests         |
| 3                | distance from shore      | 0.05            | species accessibility to kelp forests         |
| 4                | habitat occurrence       | 0.15            | species spatial overlap with kelp forests     |
| 5                | frequency of occurrence  | 0.2             | repeated kelp occurrence                      |
| 6                | life-stage               | 0.2             | ecological relevance                          |
| 7                | habitat function         | 0.3             | direct habitat use evidence                   |


***Example species***

I used a species from the Blamey and Bolton (2018) line-fish list to test the parameters and the weightings. The species is classified as "kelp-associated", commonly occurring in kelp forests with observed habitat use.

The species met the following options for the parameters:

| *Question*       | *Parameter*              | *Option*             |  *Weight*      | *alpha*        | *beta*      |
|------------------|--------------------------|----------------------|----------------|----------------|-------------|
| 1                | distribution range       | west                 | 0.05           |  4             |  1          |
| 2                | depth profile            | 15                   | 0.05           | 1.3            |  1.2        |
| 3                | distance from shore      | 3                    | 0.05           | 1.385          |  3.615      |
| 4                | habitat occurrence       | both                 | 0.15           | 6              |  1          |
| 5                | frequency of occurrence  | repeated occurrence  | 0.2            | 6              |  5          |
| 6                | life-stage               | adult                | 0.2            | 6              |  3          |
| 7                | habitat function         | reported evidence    | 0.3            | 5              |  5          |

The integrated parameters are:

$$ \alpha = (0.05 \times 4) + (0.05 \times 1.3 ) + (0.05 \times 1.385) + (0.15 \times 6) + (0.2 \times 6) + (0.2 \times 6) + (0.3 \times 5)  $$

$$ \beta = (0.05 \times 1) + (0.05 \times 1.2 ) + (0.05 \times 3.615) + (0.15 \times 1) + (0.2 \times 5) + (0.2 \times 3) + (0.3 \times 5) $$

Therefore:

$$ \theta_{H,i}^{prior} \sim Beta(5.13425	3.54075)$$

```{r}
# integrated prior

library(tidyverse)

integrate_hdr_prior <- function(question, alpha, beta, weight) {

  stopifnot(length(question) == 7)
  stopifnot(length(alpha) == 7)
  stopifnot(length(beta) == 7)
  stopifnot(length(weight) == 7)
  stopifnot(abs(sum(weight) - 1) < 0.00001)

  alpha_integrated <- sum(weight * alpha)
  beta_integrated  <- sum(weight * beta)

  tibble(
    alpha = alpha_integrated,
    beta = beta_integrated,
    mean = alpha_integrated / (alpha_integrated + beta_integrated),
    lower_95 = qbeta(0.025, alpha_integrated, beta_integrated),
    upper_95 = qbeta(0.975, alpha_integrated, beta_integrated)
  )
}

question <- c(
  "Q1 Geographic distribution",
  "Q2 Depth",
  "Q3 Distance",
  "Q4 Occurrence relative to kelp",
  "Q5 Frequency of occurrence",
  "Q6 Life-stage",
  "Q7 Habitat function"
)

alpha <- c(4, 1.3, 1.385, 6, 6, 6, 5)
beta  <- c(1, 1.2, 3.615, 1, 5, 3, 5)

weight <- c(
  0.05,
  0.05,
  0.05,
  0.15,
  0.20,
  0.20,
  0.30
)

integrated_prior <- integrate_hdr_prior(
  question = question,
  alpha = alpha,
  beta = beta,
  weight = weight
)

integrated_prior
```

Create a singular plot

```{r step2}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Figure 10: Prior beta distribution of example
#|  species habitat dependency ranking on kelp forests"
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

x <- seq(0.001, 0.999, length.out = 1000)

prior_plot_data <- tibble(
  HDR = x,
  density = dbeta(
    x,
    shape1 = integrated_prior$alpha,
    shape2 = integrated_prior$beta
  )
)

ggplot(prior_plot_data, aes(x = HDR, y = density)) +
  geom_area(fill = "#E69F00", alpha = 0.40) +
  geom_line(colour = "#CC7A00", linewidth = 1.1) +
  geom_vline(
    xintercept = integrated_prior$mean,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  annotate(
    "text",
    x = integrated_prior$mean,
    y = max(prior_plot_data$density) * 0.92,
    label = paste0(
      "Prior mean = ",
      round(integrated_prior$mean, 2)
    ),
    hjust = -0.05,
    size = 4
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +
  labs(
    title = "Integrated prior distribution for Habitat Dependency Rank",
    subtitle = paste0(
      "Beta(",
      round(integrated_prior$alpha, 2),
      ", ",
      round(integrated_prior$beta, 2),
      ")"
    ),
    x = "Habitat Dependency Rank",
    y = "Prior density"
  ) +
  theme_classic(base_size = 13)
```

