---
title: "Kelp Forest Dependence of South African Fish Species"
author: "AJ Smit, Zoë-Angelique Petersen & David Dyer"
pdf-engine: lualatex 
format:
  html: 
    df_print: paged
    toc: true 
    toc_depth: 3
    theme: journal 
    highlight: zenburn 
    number_sections: true
    code_folding: hide 
    fig_width: 7 
    fig_height: 5
---

# What is this?

This qmd in particular will be assessed to assign informative prior to assess trophic dependency ranking of line-fish on kelp forests. Currently it consists of 7 questions (parameters) in totality:

* q1-3 forms the basis of kelp accessibility (these questions will place species in regions near or outside of kelp forests, without inferences on kelp occurrence or habitat use by species)
* q4 forms the basis of a species trophic grouping, which helps infer possibilities of the consumption of kelp-based carbon sources.
* q5-7 forms the basis for use and dependency - the questions will draw on use of kelp forests, if there is a repeated, significant or exclusive use of kelp (allows conclusions of use, selection and dependency to be drawn).

The priors for this section will be set-up in two methods:
* Step 1: provides the parameters and values for Beta distributions for each parameter, which is posed as a question.

* Step 2: combines the individual parameters into a single integrated weighted prior parameter. (the weighting for each is discussed in that section and the equations are provided)

NB! This qmd will only comprise the prior set-up. The associated function code that will be used to assess species habitat priors will be in a separate qmd. This is for ease of later use and for code updates such that users can read the code and the framework alongside each other, instead of scrolling up and down to get the relevant information. 


# Trophic Dependency Ranking:

The Bayesian analysis framework aims to determine the TDR of line-fish on kelp forests. This will follow a methodological structure, of which priors will be first be set-up. The priors selected in this section will infer knowledge on the possible interactions line-fish have with kelp forests. Due to the lack of information there are with certain fish species, uncertainties will also be taken into consideration to provide a probabilistic estimate of the interaction occurring. Priors will be set up to account for this uncertainty and will later be integrated to include the different categories to display an overall TDR of line-fish on kelp forests.

## Priors

The priors are categories that will be used to determine the possible interaction between line-fish of kelp forests. These priors will encompass factors that will touch on the boundaries and extent of fish occurrence along the coast, along with the degree of consumption of kelp or kelp associated organisms. These priors were selected based on the knowledge sourced from literature sources.

The prior represents our initial *beliefs* about a parameter's effect on the TDR estimate before observing the current data. In this context, the priors will encode our expert knowledge and assumptions about the TDR based on the answers to Questions 1-x in the Survey Questionnaire. These priors will then be updated with the data (likelihood) to produce the posterior distribution, which reflects our updated beliefs about the TDR.

Each question provides qualitative information that can be quantitatively translated into prior distributions for the TDR.

### Step 1: Assigning priors selection

The priors play a big role in shaping the posterior distributions. The type of prior---whether **Diffuse**, **Weakly Informative**, or **Informative**---reflects different levels of prior knowledge or assumptions about the parameter before seeing the data. Pay careful attention to assigning priors. A diffuse or non-informative prior is chosen to exert minimal influence on the posterior distribution and allows the data to drive the posterior in the absence of strong prior beliefs. Weakly informative priors provide some guidance without strongly biasing the posterior. They reflect a minimal level of knowledge or reasonable assumptions about the parameter's likely range without assuming high confidence in specific values. Informative priors are used when there is strong prior knowledge or beliefs about the parameter.

#### **Question 1: Geographic Distribution**

Because the distribution of kelp forests and the availability of kelp as a food source vary between the west and east coasts (*Ecklonia maxima* at De Hoop) regions, fish species that are distributed in different regions along the South African coast may have different TDRs.

Select the option that best describes the geographical distribution of the fish species along the South African coast in relation to the distribution of kelp forests (*Ecklonia maxima* and/or *Laminaria pallida*). This information will help assess the species' potential TDR on kelp.

***Choices: \[West \| East \| Both\]***

-   **\[West\]**: The species is *exclusively* found in the western region of the South African (and/or Namibian) coast, which is dominated by kelp forests (*Ecklonia maxima* and/or *Laminaria pallida*). Therefore, it is *likely* that the species has a *higher HDR* on kelp, potentially even indicating an *obligate dependency* on kelp-derived carbon.
-   **\[East\]**: The species is *exclusively* found along the coastline *east of the eastern limit* of kelp forests (east of De Hoop), where *Ecklonia maxima* does not occur, and kelp forests are absent (but there's still *Ecklonia radiata*). Consequently, the species is expected to have a HDR *close to zero* (not zero due to chance of accessing *E. radiata*), indicating *no to minimal dependency* on kelp-derived carbon.
-   **\[Both\]**: The species is found in *both* the western region (with kelp forests) and the eastern region (without kelp forests). As a result, the species may have an *intermediate HDR*, suggesting a *facultative dependency* on kelp-derived carbon, where the species utilises kelp when available but can also thrive without it.

Three dependency types are accommodated by these options. **Obligate Dependency** indicates that the species relies heavily on kelp-derived carbon and may not thrive without it. **Facultative Dependency** is when species can utilise kelp-derived carbon when available but can also survive without it. Lastly, **No Dependency** when a species does not rely on kelp-derived carbon and can thrive without it.

Translation into Priors:

To model a parameter like HDR, which is a proportion bounded between 0 and 1 (or 0% to 100%), use a Beta distribution. It has a flexible shape and so will allow for various levels of certainty and mean values. The Beta distribution has two shape parameters, $\alpha$ and $\beta$, which determine the mean (@eq-equation1) and variance (@eq-equation2) of the distribution:

$$ \mu = \frac{\alpha}{\alpha + \beta} $$ {#eq-equation1}

$$ \sigma^2 = \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)} $$ {#eq-equation2} 

Example profiles of the Beta distribution for different values of $\alpha$ and $\beta$ are shown in @fig-fig1:

```{r beta_distribution}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Parameters of the Beta distribution."
#| label: fig-fig1
#| fig.align: center
#| fig-height: 9.5
#| fig-width: 12
#| fig-pos: "t"
#| fig.asp: 0.85

# Load necessary libraries
library(tidyverse)
library(ggdist)

# Define a function to calculate and plot the Beta distribution for given alpha and beta values
plot_beta_distribution <- function(alpha, beta) {
  # Create a sequence of values for the TDR (0 to 1)
  tdr_values <- seq(0, 1, length.out = 1000)
  
  # Calculate the Beta distribution density
  density_values <- dbeta(tdr_values, alpha, beta)
  
  # Calculate mean and variance
  mean_value <- alpha / (alpha + beta)
  variance_value <- (alpha * beta) / ((alpha + beta)^2 * (alpha + beta + 1))
  
  # Create a data frame for plotting
  plot_data <- data.frame(TDR = tdr_values, Density = density_values)
  
  # Create the plot
  plot <- ggplot(plot_data, aes(x = TDR, y = Density)) +
    geom_line(color = "blue", size = 1) +
    labs(title = paste("Beta Distribution (α =", alpha, ", β =", beta, ")"),
         subtitle = paste("Mean =", round(mean_value, 4), "| Variance =", round(variance_value, 4)),
         x = "Trophic Dependency Rank (TDR)",
         y = "Density") +
    theme_minimal()
  
  return(plot)
}

# Define a list of parameter combinations to explore
params <- list(
  list(alpha = 1, beta = 1),
  list(alpha = 2, beta = 5),
  list(alpha = 4, beta = 1),
  list(alpha = 1, beta = 19),
  list(alpha = 2, beta = 2),
  list(alpha = 5, beta = 5),
  list(alpha = 10, beta = 2),
  list(alpha = 5, beta = 10)
)

# Generate and display plots for each parameter combination
plots <- lapply(params, function(p) plot_beta_distribution(p$alpha, p$beta))

# Arrange the plots in a grid
library(gridExtra)
do.call(grid.arrange, c(plots, ncol = 2))
```

We can solve for α and β given the desired mean and variance.

For example, for the 'West' region, our desired mean is $\mu = 0.8$, which reflects a higher TDR. We choose a variance that indicates more uncertainty (because a species simply being confined to this region offers little certainty that it has a TDR reflective of kelp usage). Since we want a wider distribution (higher uncertainty), we can choose $\alpha$ and $\beta$ to be smaller. A Beta(4,1) distribution can be chosen to reflect a high HDR (mean = 0.8 from @eq-equation3) with moderate uncertainty (variance \~0.0267 from @eq-equation4). The calculations are as follows:

$$\mu = \frac{4}{4 + 1} = 0.8$$ {#eq-equation3}

$$\sigma^2 = \frac{4 \times 1}{(4 + 1)^2 (4 + 1 + 1)} = \frac{4}{25 \times 6} = \frac{4}{150} \approx 0.0267$$ {#eq-equation4}

The priors for the 'West,' 'East,' and 'Both' geographic regions are therefore:

| **Region** | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|------------|----------------------------|--------------|-------------------|
| **West**   | α = 4, β = 1               | μ = 0.8      | σ² ≈ 0.0267       |
| **East**   | α = 1, β = 19              | μ = 0.05     | σ² ≈ 0.00226      |
| **Both**   | α = 2, β = 2               | μ = 0.5      | σ² = 0.05         |

Visualising these priors helps us understand how they reflect our prior beliefs about the HDR. The priors are shown in @fig-fig2 and @fig-fig3:

```{r question1-priors}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Figure 2: Plot of prior distributions for the geographic 
#| distribution effect of fish regional distribution on HDR."
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the Beta distributions for the priors
priors <- data.frame(
  x = seq(0, 1, length.out = 1000),
  west = dbeta(seq(0, 1, length.out = 1000), 4, 1),
  east = dbeta(seq(0, 1, length.out = 1000), 1, 19),
  both = dbeta(seq(0, 1, length.out = 1000), 2, 2)
)

# Plot the prior distributions
ggplot(priors, aes(x = x)) +
  geom_line(aes(y = west, color = "West"), size = 1) +
  geom_line(aes(y = east, color = "East"), size = 1) +
  geom_line(aes(y = both, color = "Both"), size = 1) +
  scale_color_manual(values = c("West" = "blue", "East" = "red", "Both" = "green")) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Geographic Distribution Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.8))

```


```{r question1-priorsdis}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Figure 3: Plot of prior distributions for the geographic 
#| distribution effect of fish regional distribution on TDR."
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  West = list(alpha = 4, beta = 1),
  East = list(alpha = 1, beta = 19),
  Both = list(alpha = 2, beta = 2)
)

# Create a sequence of HDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("West", "East", "Both")))

# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Depth Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8))
```


#### **Question 2: Depth**

The priors for the depth effect on HDR will have to reflect the fact that species that occur at the same depth as kelp forests are more likely to have a higher TDR on kelp, and as we go into deeper water, the TDR on kelp is expected to decrease. Literature often reports depth as a range, broad category or maximum depth rather than consistent point estimates (depending on the assessment done). Therefore depth will be classified into categories rather than exact depth values. The depth itself does not represent occurrence in kelp, but reflects the potential for species to overlap vertically with kelp forests.

***Choices: \[Unknown \| Shallow nearshore \| Deeper nearshore \| Mid-water \| Deep water \]***

- **\[Unknown\]**: *No reliable or reported depth* parameters for species occurrences in the water column. There will be an expected *low TDR*, as unknowns have no information to suggest an overlap in kelp boundaries
- **\[Shallow nearshore\]**: Species are reported in *intertidal, surf-zone, shallow subtidal zones* with depth ranges of *0-10m* in the water column. There will be an expected *moderately high TDR* as there is a stronger opportunity for a possible overlap with canopy kelp systems.
- **\[Deeper nearshore\]**: Species are reported in depths from approximately *10-30m* in the water column. There is an expected *moderately high TDR*, as there is an opportunity for possible overlap with kelp forest systems.
- **\[Mid-water\]**: Species are reported mainly in offshore reefs, shelf waters or mid-water habitats that range in depth from *30-60m*. There is an expected *moderately low TDR*, as there is a limited potential of overlap, with possible deeper kelp forest systems which are not common. and episodic movements between vertical columns.   
- **\[Deep water\]**: Species has a reported occurrence in deeper waters from which are greater than 60m. There is an expected *very low TDR* as there is a very low potential for kelp overlap.
- **\[Many\]**: Species are reported at greater depth ranges encompassing depths from two or more options. There is an expected *moderately low TDR* as species occur outside of ~E. maxima~ distribution boundaries, while occurring in depths suitable for ~E. radiata~  

NB! In kelp boundaries use the shallowest documented ecological relevant depth rather than the maximum depth. If a species occurs in shallow waters but also deeper waters but along the 'West' coast i.e. 5-150m, then report shallow option. For species outside of kelp boundaries use 'Many'.

*Step 1: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* Unknown ~ 0.5 (low TDR)
* Shallow nearshore ~ 0.7 (very high TDR)
* Deep nearshore ~ 0.6 (high DR)
* Mid-water ~ 0.35 (moderately low TDR)
* Deep water ~  0.125 (low TDR)
* Many ~ 0.3 (moderate TDR)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for parameter combinations:

1. Unknown (target mean = 0.5)
Using mean = α/(α+β) = 0.5


$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5 α + 0.5 β
$$\

$$
0.5α = 0.5 β
$$\

$$
α = β
$$\

I selected β = 1 and giving α = 1\ (uninformative prior)

2. Shallow nearshore (target mean = 0.7):
Using mean = α/(α+β) = 0.7


$$
\frac{α}{(α+β)} = 0.7
$$\

$$
α = 0.7 \cdot (α+β)
$$\

$$
α = 0.7α + 0.7β
$$\

$$
0.3α = 0.7β
$$\

$$
α = \frac{7}{3}β
$$\

I selected β = 3 and α = 7\

3. Deeper nearshore (target mean = 0.6):

* Using similar equations: α/(α+β) = 0.6

$$
\frac{α}{(α+β)} = 0.6
$$\

$$
α = 0.6 \cdot (α+β)
$$\

$$
α = 0.6α + 0.6β
$$\

$$
0.4α = 0.6β
$$\

$$
α = 1.5β
$$\

I selected β = 4 and α = 6

4. Mid-water (target mean = 0.3):

* Using mean = α/(α+β) = 0.3

$$
\frac{α}{(α+β)} = 0.3
$$\

$$
α = 0.3 \cdot (α+β)
$$\

$$
α = 0.3α + 0.3β
$$\

$$
0.7α = 0.3β
$$\

$$
α = 0.429β
$$\

I selected β = 7 and α = 3

5. Deep water (target mean = 0.125):

* Using mean = α/(α+β) = 0.125

$$
\frac{α}{(α+β)} = 0.125
$$\

$$
α = 0.125 \cdot (α+β)
$$\

$$
α = 0.125α + 0.125β
$$\

$$
0.875α = 0.125β
$$\

$$
α = 0.143β
$$\

I selected β = 7 and α = 1

6. Many (target mean = 0.35):
This represents a very strong prior belief that the dependency score is close to one

* Using mean = α/(α+β) = 0.35

$$
\frac{α}{(α+β)} = 0.35
$$\

$$
α = 0.35 \cdot (α+β)
$$\

$$
α = 0.35α + 0.35β
$$\

$$
0.65α = 0.35β
$$\

$$
α = 0.538β
$$\

I selected β = 6 and α = 3.228


Translated into Priors:

| **Depth**               | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|-------------------------|----------------------------|--------------|-------------------|
| **Unknown**             | α = 1, β = 7               | μ = 0.5      | σ² ≈ 0.0833       |
| **Shallow nearshore**   | α = 7, β = 3               | μ = 0.7      | σ² ≈ 0.0191       |
| **Deeper nearshore**    | α = 6, β = 4               | μ = 0.6      | σ² ≈ 0.0218       |
| **Mid-water**           | α = 3, β = 7               | μ = 0.3      | σ² ≈ 0.0191       |
| **Deep water**          | α = 1, β = 7               | μ = 0.125    | σ² ≈ 0.0122       |
| **Many**                | α = 3.228, β = 6           | μ = 0.35     | σ² ≈ 0.0222       |

A visualisation of these priors is given in @fig-fig4:

```{r question2-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of species depth profiles along the South African coast."
#| label: fig-fig4
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  #Unknown = list(alpha = 1, beta = 1), (uninformative prior)
  Shallow_nearshore = list(alpha = 7, beta = 3),
  Deeper_nearshore = list(alpha = 6, beta = 4),
  Mid_water = list(alpha = 3, beta = 7),
  Deeper_water = list(alpha = 1, beta = 7),
  Many = list(alpha = 3.228, beta = 6)
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(HDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Kelp-use",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```

#### **Question 3: Coastal position**

The priors for the distance from the shore effect on TDR will have to reflect the fact that species that occur closer to the shore are more likely to have a higher HDR on kelp, and as we move further offshore, the TDR on kelp is expected to decrease. The priors for this question are as follows:

***Choices: \[ Unknown \|Immediate nearshore \| Nearshore-offshore continuum \| Offshore \| Pelagic \]*** 

- **\[Unknown\]**: There is *no clear coastal position* described for species in South African waters. There is an expected *low TDR* as no inferences can be drawn about a species accessibility to kelp forests.
- **\[Immediare nearshore\]**: Species occupies zones *closer to the shore* such as intertidal, surf-zone, tidal pool, estuary mouth, inshore, nearshore, shallow coastal reef or rocky shore. There will be an expected *moderately high TDR* as these zones offer the highest opportunity for species accessibility to kelp forests.
- **\[Offshore\]**: The species occurs in habitats *further from the shore* which include zones noted as offshore reefs, shelf waters, outer coastal zones, continental shelf, deeper coastal waters. There is an expected *low TDR* as there is a limited opportunity for species to access kelp forest systems.
- **\[Nearshore-offshore\]**: Species occupies zones in between nearshore and offshore regions, together with both zones. These zones include habitats such as inshore and offshore reefs, coastal waters, bays and embayments, reef-associated migrations can be included in this section. There is an expected *moderate TDR* as there is a possible opportunity for species to access kelp forest environments.  
- **\[Pelagic\]**: Species occupies zones that are *off the continental shelf*. These include zones such as pelagic, epipelagic, oceanic, open water column and offshore water. There is an expected *low TDR* as species has the lowest opportunity to access kelp forests.
- **\[Many\]**: Species occupies nearshore and/or offshore zones together with zones that are *off the continental shelf*. There is an expected *moderately low TDR* as species has a low opportunity to access kelp forests.

Translated into Priors:

| **Coastal position**      | **Beta Parameters (α, β)** | **Mean (μ)** | **Variance (σ²)** |
|---------------------------|----------------------------|--------------|-------------------|
| **Unknown**               | α = 1, β = 1               | μ ≈ 0.5      | σ² ≈ 0.0833       |
| **Immediate nearshore**   | α = 7, β = 3               | μ = 0.7      | σ² ≈ 0.0191       |
| **Offshore**              | α = 3, β = 7               | μ ≈ 0.3      | σ² ≈ 0.0191       |
| **Nearshore-offshore**    | α = 5, β = 5               | μ ≈ 0.5      | σ² ≈ 0.0227       |
| **Pelagic**               | α = 1, β = 7               | μ ≈ 0.125    | σ² ≈ 0.0122       |
| **Many**                | α = 3.228, β = 6           | μ = 0.35     | σ² ≈ 0.0222       |

A visualisation of these priors is given in @fig-fig5:

```{r question3-prior}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of species coastal position along the South African coast."
#| label: fig-fig5
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  #Unknown = list(alpha = 1, beta = 1), (uninformative prior)
  Immediate_nearshore = list(alpha = 7, beta = 3),
  Offshore = list(alpha = 3, beta = 7),
  Nearshore_offshore = list(alpha = 5, beta = 5),
  Pelagic = list(alpha = 1, beta = 7),
  Many = list(alpha = 3.228, beta = 6)
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Kelp-use",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```


#### **Question 4: Trophic hierarchy**\

1.  Define Primary Categories for Trophic Level Start with three main trophic level categories:\

* 𝜃~Herbivore~: Probability of a species being strictly herbivorous. 
* 𝜃~Carnivore~: Probability of a species being strictly carnivorous. 
* 𝜃~Omnivore~: Probability of a species being omnivorous.

Define these in a Dirichlet distribution:

$$
(𝜃_{Herbivore} , 𝜃_{Carnivore} , 𝜃_{Omnivore}) ∼ Dirichlet (𝛼_{Herbivore} ,
𝛼_{Carnivore} , 𝛼_{Omnivore})
$$\

2.  Establish Proportions within the Omnivore Category\

Since an omnivore may have a mixed diet of animal and plant-based items, let's introduce two sub-parameters within the omnivore category to capture the dietary proportions: 

* 𝜙~Plant~: Proportion of the omnivore diet that is plant-based. 
* 𝜙~Animal~: Proportion of the omnivore diet that is animal-based.

These should satisfy:\

$$
𝜙_{Plant} + 𝜙_{Animal} = 1
$$\

To account for different dietary proportions, model 𝜙~Plant~ and 𝜙~Animal~ with a Dirichlet prior as well:\

$$
(𝜙_{Plant} , 𝜙_{Animal}) ∼ Dirichlet (𝛽_{Plant} , 𝛽_{Animal})
$$\

where 𝛽~Plant~ and 𝛽~Animal~ are the concentration parameters that reflect prior beliefs about the typical proportion of plant and animal components in an omnivorous diet.\

3. Integrate the Omnivore Proportions into the Overall Model\

Now,if the overall probability of an omnivorous diet is 𝜃~Omnivore~, then the expected contribution of plant and animal items to the omnivorous diet is given by:\

$$
𝜃_{Omnivore, Plant} = 𝜃_{Omnivore} × 𝜙_{Plant} 
$$\
$$
𝜃_{Omnivore, Animal} = 𝜃_{Omnivore} × 𝜙_{Animal} 
$$\

4. Define the Joint Model\

The joint model now includes both the trophic level and dietary proportions: 
• First, assign a Dirichlet distribution to the main trophic levels (𝜃~Herbivore~ ,𝜃~Carnivore~ , 𝜃~Omnivore~ ). 
• Next, conditionally on the species being omnivorous, assign a Dirichlet distribution to the proportions of plant and animal components within the omnivore category. 

The final model reflects both the probability of a species being omnivorous and the specific proportions within that dietary choice:\

$$
𝜃 =(𝜃_{Herbivore} , 𝜃_{Carnivore} , 𝜃_{Omnivore, Plant} , 𝜃_{Omnivore,Animal} ) ∼ Dirichlet(𝛼_{Herbivore}, 𝛼_{Carnivore} , 𝛼_{Omnivore} ⋅𝜙_{Plant} , 𝛼_{Omnivore} ⋅𝜙_{Animal})
$$

5. Choosing Concentration Parameters\

The concentration parameters 𝛼 and 𝛽 determine the strength of prior belief in each category. If 𝜙~Plant~ = 0.8 and 𝜙~Animal~ = 0.2, then the parameters 𝛽~Plant~ = 8 and 𝛽~Animal~ = 2 reflect moderate certainty in this distribution.Adjust 𝛼Omnivore accordingly based on the observed data or expert input.

***Process for selecting Alpha and Beta Paramaters***\

*Step 1: Understanding the Ecological Relationship*\
The expected TDR of line fish will be high for individuals that are herbivores, and smaller for line fish that are carnivores. Omnivores display a mixed diet, and will thus have an intermediate TDR.\

The TDR of omnivores will differ based on the proportion of plant and animal consumed by line fish. Omnivores with a greater proportion of plants to animals in their diet will display a greater TDR, than those with a greater proportion of animals to plants in their diet.\

*Step 2: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* Herbivore ~ 0.8 (high dependency)
* Omnivore ~ 0.55 (moderate dependency)
* Carnivores ~ 0.25 (low dependency)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. Herbivores (target mean = 0.8)
Using mean = α/(α+β) = 0.8

$$
\frac{α}{(α+β)} = 0.8
$$\

$$
α = 0.8 \cdot (α+β)
$$\

$$
α = 0.8α + 0.8β
$$\

$$
0.2α = 0.8β
$$\

$$
α = 4β
$$\

I selected β = 2 for moderate variance, giving α = 8\

2. Omnivores (target mean = 0.556):

* Using similar equations: α/(α+β) = 0.556

$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5α + 0.5β
$$\

$$
0.5α = 0.5β
$$\

$$
α = 1β
$$\

I selected β = 5, giving α = 5

3. Carnivores (target mean = 0.25):

* Using mean = α/(α+β) = 0.25

$$
\frac{α}{(α+β)} = 0.25
$$\

$$
α = 0.25 \cdot (α+β)
$$\

$$
α = 0.25α + 0.25β
$$\

$$
0.75α = 0.25β
$$\

$$
α = \frac{1}{3}β
$$\

I selected β = 6, giving α = 2

*Step 5: Verifying Distribution Shapes*\
For each parameter set, I verified:

* The shape matched ecological expectations
* There was appropriate separation between distributions
* Variance was reasonable (not too narrow or wide)

*Step 6: Checking Practical Interpretation*\
Finally, I confirmed that:

* The distributions aligned with ecological understanding
* The overlaps between categories reflected natural variation
* The mode and concentration were appropriate for biological variability in trophic dependencies

***Visual representation of priors***\

|**Trophic hierarchy** | **Mean TDR** | **α** | **β** |
|------------------|--------------|-------|-------|
|**Herbivore**     | 0.8          | 8     | 2     | 
|**Omnivore**      | 0.557        | 5     | 4     |
|**Carnivore**     | 0.25         | 2     | 6     |


Plot the Beta prior distributions @fig-fig6:

```{r}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distributions of species trophic groupings"
#| label: fig-fig6
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65


# Load packages
library(tidyverse)

# Define the parameters for each option
options <- list(
  "Herbivore" = list(alpha = 8, beta = 2),
  "Omnivore" = list(alpha = 5, beta = 5),
  "Carnivore" = list(alpha = 2, beta = 6)
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("Herbivore", "Omnivore",
                                            "Carnivore")))

# Plot the densities
class(density_long$Option)


ggplot(density_long, aes(x = TDR, y = Density, colour = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Trophic Mode Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.92),
        legend.text = element_text(size = 11))

```

#### **Question 5: Kelp/kelp associated prey**

In this question you are asked to select the answer that best described the occurrence of kelp or kelp-associated prey items (e.g. Haliotis midae, Parenchinus angulosus, Jasus lalandi) in their stomachs contents across habitats. Answering 'Yes' or 'Disproportionate' for instances where there was prey items in the species diet. 'None' will be used when no prey items were found in the species diet.  

***Choices: \[None \| Yes\| Disproportionate \]***

- **\[None\]**: The species diet comprised *no kelp or kelp-associated prey items*. There is an expected *Low TDR* as there is *no evidence* supporting species use of kelp through a trophic route. 
- **\[Yes\]**: The species diet comprised *kelp or kelp-associated prey items*. There is an expected *moderately High TDR*, as there is evidence of kelp use through a trophic route.
- **\[Disproportionate\]**: The species diet comprise a *disproportionate amount of kelp or kelp-associated prey items* relative to other dietary items. There is an expected *High TDR* as the species indicates a selection for kelp or kelp-associated prey items, which supports a strong trophic route.

*Step 1: Understanding the Ecological Relationship*\
The expected TDR of line fish will be high for individuals that comprise a disproportionate amount of kelp or kelp-associated prey items, and smaller for line fish that have no kelp or kelp-associated prey items in their diet. Species that comprise kelp/kelp associated prey comprise a variety of other organisms that may or may not be linked to kelp, thus a moderate TDR is expected.\

The TDR of yes will differ based on the proportion kelp present. Species with a greater proportion of kelp or kelp-associated prey items will display a greater TDR, than those with a greater proportion non-kelp linked prey items in their diet.\

*Step 2: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* Disproportionate ~ 0.8 (high dependency)
* Yes ~ 0.5 (moderate dependency)
* No ~ 0.2 (low dependency)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. Disproportionate (target mean = 0.80)
Using mean = α/(α+β) = 0.8

1. Disproportionate (target mean = 0.8)
Using mean = α/(α+β) = 0.8

$$
\frac{α}{(α+β)} = 0.8
$$\

$$
α = 0.8 \cdot (α+β)
$$\

$$
α = 0.8α + 0.8β
$$\

$$
0.2α = 0.8β
$$\

$$
α = 4β
$$\

I selected β = 2 and α = 8, giving a narrow variance, as this is a highly informative prior\

2. Yes (target mean = 0.65):

* Using similar equations: α/(α+β) = 0.5

$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5α + 0.5β
$$\

$$
0.5α = 0.5β
$$\

$$
α = 1β
$$\

I selected β = 5 and α = 5, giving a moderate variance as kelp or kelp-associated prey items in species diet does not indicate if species actively select those prey items. I.e. species can accidentally or incidentally consume these prey items.

3. None (target mean = 0.11):

* Using mean = α/(α+β) = 0.11

$$
\frac{α}{(α+β)} = 0.11
$$\

$$
α = 0.11 \cdot (α+β)
$$\

$$
α = 0.11α + 0.11β
$$\

$$
0.89α = 0.11β
$$\

$$
α = \frac{89}{11}β
$$\

I selected β = 9 and α = 1.1, giving a narrow variance. I wasn't sure between a bell shape or an immediate curve ()


Translated into priors:

| **Kelp/kelp-associated prey** | **Mean TDR** | **α** | **β** |
|-------------------------------|--------------|-------|-------|
|**None**                       | 0.11         | 1.1   | 9     | 
|**Yes**                        | 0.5          | 5     | 5     |
|**Disproportionate**           | 0.8          | 8     | 2     |


A visualisation of these priors is given in @fig-fig7:

```{r question5-priordis}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distribution of TDR based on species consumption of kelp or kelp-associated prey items"
#| label: fig-fig7
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65


# Define the parameters for each option
options <- list(
  "Disproportionate" = list(alpha = 8, beta = 2),
  "Yes" = list(alpha = 5, beta = 5),
  "No" = list(alpha = 1.1, beta = 9)
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("Disproportionate", 
"Yes", "No")))

# Plot the densities
class(density_long$Option)


ggplot(density_long, aes(x = TDR, y = Density, colour = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Consumption of kelp/kelp-associated prey",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.80),
        legend.text = element_text(size = 11))

```

#### **Question 6: Proportion of kelp/kelp associated organisms consumed** 

The proportion of stomachs that kelp or kelp associated organisms were found in can indicate the degree at which species rely on those particular food types. The frequency of occurrence (FO) indicate the number of stomachs that comprise kelp or kelp associated species. Species with a greater FO of food items that are kelp/associated to kelp are more likely to have a higher TDR on kelp, while as the FO of food items consumed decreases, the TDR on kelp is expected to decrease. The priors for this section are as follows:

***Choices: \[None \| Low\| Moderate \| High \| Greatest\]***

- **\[None\]**: The species has *no kelp or kelp associated prey items* defined by *FO of 0* in stomachs. There is an expected *low TDR*, as species display no kelp-link.
- **\[Low\]**: The species has a *low proportion* defined by *FO ranging between 0-0.4* of kelp or kelp-associated prey items in their dietary composition. There is an expected *low TDR*, as species display a minimal kelp-link.
- **\[Moderate\]**: The species has a *moderate proportion* defined by a *FO ranging between 0.4-0.6* of kelp or kelp associated prey items in stomachs. There is an expected *moderate TDR*, as species display a moderate kelp-link.  
- **\[High\]**: The species has a *high proportion* defined by a *FO ranging between 0.6-0.8* of kelp or kelp-associated prey items in stomachs. There is an expected *high TDR*, as species display a high kelp-link.
- **\[Greatest\]**: The species has the *greatest proportion* defined by a *FO >0.8* of kelp or kelp-associated prey items in stomachs. There is an expected *very high TDR*, as species display a near-exclusive dietary composition comprising kelp-based carbon sources.

Translate these priors into Beta distributions:

```{r}
# for the mean
(2)/(2+8)

# for the variance
(2*8) / ((2+8)^2 * (2+8+1))


# from claude

# Define prior parameters for different dependency levels
prior_params <- list(
  greatest_proportion = list(
    alpha = 19, beta = 1,    # Mean ~0.8, concentrated distribution
    description = "Greatest proportion of kelp/kelp-associated items"
  ),
  high_proportion = list(
    alpha = 15, beta = 5,
    description = "High proportion of kelp/kelp associated items"
  ),
  moderate_proportion = list(
    alpha = 5, beta = 5,    # Mean 0.5, moderate variance
    description = "Mixed diet with substantial kelp component"
  ),
  low_proportion = list(
    alpha = 4, beta = 10,
    description = "Low proportion of kelp items with mixed diet"
  ),
  lowest_proportion = list(
    alpha = 2, beta = 8,    # Mean ~0.2, indicates occasional kelp consumption
    description = "Lowest proportion of kelp items"
  )
)

# Function to calculate mean and variance
get_distribution_stats <- function(alpha, beta) {
  mean = alpha / (alpha + beta)
  variance = (alpha * beta) / ((alpha + beta)^2 * (alpha + beta + 1))
  return(list(mean = mean, variance = variance))
}

# Calculate metrics for each prior
for(dep_level in names(prior_params)) {
  stats <- get_distribution_stats(
    prior_params[[dep_level]]$alpha, 
    prior_params[[dep_level]]$beta
  )
  cat(sprintf(
    "%s:\n  Mean: %.2f\n  Variance: %.3f\n\n",
    dep_level, stats$mean, stats$variance
  ))
} 
```

| **Proportion**    | **alpha**  |  **beta**  | **mean**  | **variance**  |
|-------------------|------------|------------|-----------|---------------|
| **\[None\]**      | α = 1      | β = 19     | 0.05      | 0.015         |
| **\[Low\]**       | α = 2      | β = 15     | 0.29      | 0.026         |
| **\[Moderate\]**  | α = 10     | β = 10     | 0.5       | 0.012         |
| **\[High\]**      | α = 15     | β = 5      | 0.75      | 0.009         |
| **\[Greatest\]**  | α = 19     | β = 1      | 0.95      | 0.002         |


Now lets visualize the Beta distributions for diet proportion @fig-fig8:

```{r}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distributions of frequency of occurrence of kelp or kelp-associated prey items in species stomachs"
#| label: fig-fig8
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

library(tidyverse)
# Define the scaling factor 
# scale_factor <- 4.55  # Adjust this value as needed

# Define the parameters for each option
options <- list(
  "Greatest_proportion" = list(alpha = 19, beta = 1),
  "High_proportion" = list(alpha = 15, beta = 5),
  "Moderate_proportion" = list(alpha = 5, beta = 5),
  "Low_proportion" = list(alpha = 4, beta = 10),
  "None" = list(alpha = 1, beta = 19) 
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("Greatest_proportion", 
                                            "High_proportion", 
                                            "Moderate_proportion", 
                                            "Low_proportion",
                                            "None")))



# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Proportion of Consumption Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8),
        legend.key.size = unit(0.5, 'cm'),
        legend.text = element_text(size = 11))
```


#### **Question 7: Trophic function**

In this question you are asked to select the answer that best describes whether kelp or kelp-associated prey items are functionally important in the diet of line-fish. The options 'Common', 'Significant' or 'Dependent' applies to instances where you answered 'Yes' and 'Disproportionate' in question 5 and any option > '0' in question 6. The none option applies when there is kelp is not detected to have a linkage to the trophic functioning of a species.  

***Choices: \[None \| Common \| Significant \| Dependent \]***

- **\[None\]**: The species diet comprised *no kelp or kelp-associated prey items*. There is an expected *Low TDR* as there is *no evidence* supporting species use of kelp through a trophic route. 
- **\[Common\]**: The species diet *commonly comprised kelp or kelp-associated prey items*. There is an expected *moderately High TDR*, as there is evidence of kelp use through a trophic route.
- **\[Significant\]**: The species diet comprise a *significant occurrence of kelp or kelp-associated prey items* relative to other dietary items, improving a specie relative fitness. There is an expected *High TDR* as the species indicates a selection for kelp or kelp-associated prey items, which supports a strong trophic route.
- **\[Dependent\]**: The species require a diet of kelp or kelp-associated prey items to allow for a continued life-span. Without kelp or kelp-associated species, the species will undergo a population decline whereby alternative habitats cannot replace the function lost from kelp forests. There is an *expected high TDR* as the species requires kelp to survive.


*Step 1: Understanding the Ecological Relationship*\
The expected TDR of line fish will be high for individuals that have a greater kelp trophic function and smaller for line fish that have no kelp trophic function. 

- **\[None\]**: No or currently tested kelp or kelp-associated prey link. This places the species trophic pathway in alternative ecosystems. 
- **\[Common\]**: Kelp or kelp-associated prey items occur repeatedly in their stomachs but which form part one component of a broad diet. The trophic functioning is moderate, as kelp forms only a part of their function.
- **\[Significant\]**: Kelp or kelp-associated prey items are disproportionately represented relative to alternative food sources. It indicates a selection of kelp-based carbon sources.
- **\[Dependent\]**: There is strong evidence and source justification for an ecological improvement in line-fish ecological fitness. Without kelp forests present, there would be a decline in species abundance, whereby alternative ecosystems cannot replace the trophic pathway. 

*Step 2: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* None ~ 0.12 (low dependency)
* Common ~ 0.54 (moderate dependency)
* Significance ~ 0.75 (high dependency)
* Dependent ~ 0.9 (very high dependency)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. None (target mean = 0.12)
Using mean = α/(α+β) = 0.12


$$
\frac{α}{(α+β)} = 0.12
$$\

$$
α = 0.12 \cdot (α+β)
$$\

$$
α = 0.12α + 0.12β
$$\

$$
0.88α = 0.12β
$$\

$$
α = \frac{0.12}β
$$\

I selected β = 8.8 for narrow variance, giving α $\approx$ 0.12\

2. Common (target mean = 0.54):

* Using similar equations: α/(α+β) = 0.54

$$
\frac{α}{(α+β)} = 0.54
$$\

$$
α = 0.54 \cdot (α+β)
$$\

$$
α = 0.54α + 0.54β
$$\

$$
0.46α = 0.54β
$$\

$$
α = \frac{27}{23}β
$$\

I selected β = 4.6 and α = 5.4

3. Significant (target mean = 0.75):

* Using mean = α/(α+β) = 0.75

$$
\frac{α}{(α+β)} = 0.75
$$\

$$
α = 0.15 \cdot (α+β)
$$\

$$
α = 0.75α + 0.75β
$$\

$$
0.25α = 0.75β
$$\

$$
α = \frac{3}β
$$\

I selected β = 2.5 and α = 7.2

4. Dependence (target mean = 0.9):

* Using mean = α/(α+β) = 0.9

$$
\frac{α}{(α+β)} = 0.9
$$\

$$
α = 0.9 \cdot (α+β)
$$\

$$
α = 0.9α + 0.9β
$$\

$$
0.1α = 0.9β
$$\

$$
α = \frac{9}β
$$\

I selected β = 1 and α = 9

Translated into priors:

| **Trophic function** | **Mean TDR** | **α** | **β** |
|----------------------|--------------|-------|-------|
|**None**              | 0.12         | 1.2   | 8.8   | 
|**Common**            | 0.54         | 5.4   | 4.6   |
|**Significant**       | 0.75         | 7.5   | 2.5   |
|**Dependent**         | 0.90         | 9     | 1     |


A visualisation of these priors is given in @fig-fig9:

```{r}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior Beta distributions of reported trophic function of species in kelp forests"
#| label: fig-fig9
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

# Define the parameters for each option
options <- list(
  None = list(alpha = 1.2, beta = 8.8),
  Common = list(alpha = 5.4, beta = 4.6),
  Significant = list(alpha = 7.5, beta = 2.5),
  Dependent = list(alpha = 9, beta = 1)
)

# Create a sequence of HDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Trophic function",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.4, 0.8))
```

### Step 2: Integrating the prior parameters

Since priors play an important role in shaping posteriors, I wanted to ensure prior Beta distributions for each parameter was carefully generated. These parameters are not single entities, but rather form collective insights of how line-fish use kelp forests. The first three questions (Q1-3) form part of the baseline assessment that places a species in kelp forest boundaries. Questions 4-6 provides insight into the use of kelp forests by line-fish, whether the occurrence of prey items in the diet of species is mere incidental occurrence, or if a trophic use. Question 7 provides greater clarity on kelp forest use: whether there would be a species population decline in response to kelp loss. Thus, each of these parameters hold different weights in the final outcome:\

#### Weights

As mentioned, each section will weigh differently in the final integration. The weights will add up to 1. 

- Since q1-3 forms the baseline of the assessment, placing a species in kelp boundaries, the weights of these will be low as they do not inform strong relationship but rather a possibility. They will be weighted low.
- Q4 places a species into their respective trophic groupings, it describes their dominant feeding group and prey items found in their stomachs. This parameter will have a greater weight than q1-3 but not as high as it does not inform kelp trophic use, rather species feeding biology.
- Q5 places infers whether there is a dietary composition of kelp or kelp-associated prey items in the stomachs of line fish. As it goes a bit deeper into the trophic use of line-fish in kelp forest. It will weigh more than q1-4.
- Q6 provides information on species species prey item proportion, a greater detail into the frequency of occurrences of prey items. It provides with a greater clarity, if kelp-based carbon sources occur more readily than alternative prey items in stomachs. It will have a greater weight than q1-5, due to the detail that it provides.
- Q7 will provide insights into trophic functioning by species, indicating whether if kelp-based carbon sources are incidental occurrence, if there is an ecological use, selection (based on disproportionate occurrences), or whether they are dependent on kelp forests to under a continuation their life-cycle. This will have the greatest weight as it places the greatest insight into habitat use by line-fish.

#### How to read the Beta distributions?

* values of 1 indicating dependency on kelp forests by line-fish\
* values above 0.5 indicating a strong use of kelp\
* values below 0.5 indicating a weak kelp use\
* and 0 indicating no relationship to kelp forests. 

Thus, values closer to 1 on the beta distribution indicates a stronger relationship between line-fish and kelp.

In order to assess this, a weighted logarithmic pool of seven beta distributions will be used:

Weight log pool:
$$ \pi_{TDR,i} (\theta) \propto \prod_{q=1}^{7} \left[ \pi_q(\theta \mid \alpha_{qi}, \beta_{qi}) \right] w_q $$


where w~q~ $\ge$ 0 and:

$$ \sum_{q=1}^{7} w_q = 1 $$

For Beta distributions, this remains a beta distribution:

$$ \theta_{H,i}^{prior} \sim Beta(\alpha_i, \beta_i) $$

with:

$$ \alpha_i = 1 +\sum_{q=1}^{7} w_q(\alpha_{qi} - 1) $$

$$ \beta_i = 1 +\sum_{q=1}^{7} w_q(\beta_{qi} - 1) $$

because the weights sum to 1, this is equivalently:

$$ \alpha_i = \sum_{q=1}^{7} w_q \alpha_{qi} $$

$$ \beta_i = \sum_{q=1}^{7} w_q \beta_{qi} $$

Altogether, this generates a single integrated prior, whereby weighted pooling combines distributions using weighted mean and retains a Beta form.

Possible weights for each question:

| *Question*       | *Parameter*              | *Weight*        | *Role*                                        |
|------------------|--------------------------|-----------------|-----------------------------------------------|
| 1                | Distribution range       | 0.05            | species accessibility to kelp forests         |
| 2                | Depth profile            | 0.05            | species accessibility to kelp forests         |
| 3                | Distance from shore      | 0.05            | species accessibility to kelp forests         |
| 4                | Trophic hierarchy        | 0.10            | species spatial overlap with kelp forests     |
| 5                | Kelp-based sources       | 0.15             | repeated kelp occurrence                      |
| 6                | Proportion of prey items | 0.2             | ecological relevance                          |
| 7                | Trophic function         | 0.4             | direct habitat use evidence                   |


***Example species***

I used a species from the Blamey and Bolton (2018) line-fish list to test the parameters and the weightings. The species is classified as "kelp-associated", commonly occurring in kelp forests with observed habitat use.

The species met the following options for the parameters:

| *Question*       | *Parameter*              | *Option*             |  *Weight*      | *alpha*        | *beta*      |
|------------------|--------------------------|----------------------|----------------|----------------|-------------|
| 1                | Distribution range       | west                 | 0.05           | 4              |  1          |
| 2                | Depth profile            | 15                   | 0.05           | 1.3            |  1.2        |
| 3                | Distance from shore      | 3                    | 0.05           | 1.385          |  3.615      |
| 4                | Trophic hierarchy        | Omnivore             | 0.10           | 5              |  4          |
| 5                | Kelp-based sources       | Yes                  | 0.15           | 5              |  5          |
| 6                | Proportion of prey items | Low                  | 0.2            | 2              |  15         |
| 7                | Trophic function         | Common               | 0.4            | 5.4            |  4.6        |

The integrated parameters are:

$$ \alpha = (0.05 \times 4) + (0.05 \times 1.3 ) + (0.05 \times 1.385) + (0.10 \times 5) + (0.15 \times 5) + (0.2 \times 2) + (0.4 \times 5.4)  $$

$$ \beta = (0.05 \times 1) + (0.05 \times 1.2 ) + (0.05 \times 3.615) + (0.10 \times 14 + (0.15 \times 5) + (0.2 \times 15) + (0.4 \times 4.6) $$

Therefore:

$$ \theta_{H,i}^{prior} \sim Beta(4.14425,	6.28075)$$

```{r}
# integrated prior

library(tidyverse)

integrate_hdr_prior <- function(question, alpha, beta, weight) {

  stopifnot(length(question) == 7)
  stopifnot(length(alpha) == 7)
  stopifnot(length(beta) == 7)
  stopifnot(length(weight) == 7)
  stopifnot(abs(sum(weight) - 1) < 0.00001)

  alpha_integrated <- sum(weight * alpha)
  beta_integrated  <- sum(weight * beta)

  tibble(
    alpha = alpha_integrated,
    beta = beta_integrated,
    mean = alpha_integrated / (alpha_integrated + beta_integrated),
    lower_95 = qbeta(0.025, alpha_integrated, beta_integrated),
    upper_95 = qbeta(0.975, alpha_integrated, beta_integrated)
  )
}

question <- c(
  "Q1 Geographic distribution",
  "Q2 Depth",
  "Q3 Distance",
  "Q4 Trophic hierarchy",
  "Q5 Kelp-based carbon source",
  "Q6 Proportion of prey items",
  "Q7 Trophic function"
)

alpha <- c(4, 1.3, 1.385, 5, 5, 2, 5.4)
beta  <- c(1, 1.2, 3.615, 4, 5, 15, 4.6)

weight <- c(
  0.05,
  0.05,
  0.05,
  0.10,
  0.15,
  0.20,
  0.40
)

integrated_prior <- integrate_hdr_prior(
  question = question,
  alpha = alpha,
  beta = beta,
  weight = weight
)

integrated_prior
```

Create a singular plot @fig-fig10:

```{r step2}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Prior beta distribution of example
#|  species trophic dependency ranking on kelp forests"
#|  label: fig-fig10
#| #| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65

x <- seq(0.001, 0.999, length.out = 1000)

prior_plot_data <- tibble(
  HDR = x,
  density = dbeta(
    x,
    shape1 = integrated_prior$alpha,
    shape2 = integrated_prior$beta
  )
)

ggplot(prior_plot_data, aes(x = HDR, y = density)) +
  geom_area(fill = "#E69F00", alpha = 0.40) +
  geom_line(colour = "#CC7A00", linewidth = 1.1) +
  geom_vline(
    xintercept = integrated_prior$mean,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  annotate(
    "text",
    x = integrated_prior$mean,
    y = max(prior_plot_data$density) * 0.92,
    label = paste0(
      "Prior mean = ",
      round(integrated_prior$mean, 2)
    ),
    hjust = -0.05,
    size = 4
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +
  labs(
    title = "Integrated prior distribution for Trophic Dependency Rank",
    subtitle = paste0(
      "Beta(",
      round(integrated_prior$alpha, 2),
      ", ",
      round(integrated_prior$beta, 2),
      ")"
    ),
    x = "Trophic Dependency Rank",
    y = "Prior density"
  ) +
  theme_classic(base_size = 13)
```