---
title: "Kelp Forest Dependence of South African Fish Species"
author: "AJ Smit, Zoë-Angelique Petersen & David Dyer"
pdf-engine: lualatex 
format:
  html: 
    df_print: paged
    toc: true 
    toc_depth: 3
    theme: journal 
    highlight: zenburn 
    number_sections: true
    code_folding: hide 
    fig_width: 7 
    fig_height: 5
---

# Bayesian Analysis Framework

The Bayesian analysis framework underpins the assessment of clarifying the dependency of line-fish into categories of dependency or association based on their relationship with kelp forests. Previous literature studies have been nuanced in their descriptions and understanding with the terms dependency and association, often using these terms synonymously. The background of this models aims to establish a difference between the terms dependency and association with each being specified in accordance with the following belong:\

-   *Ecological benefit*: Increased fitness
-   *Association*: The geographical overlap of line-fish and kelp forests along the coast, with no ecological benefit derived from kelp forests by line-fish
-   *Dependency*: Line-fish obtain an ecological benefit from kelp forests, either in the form of habitat or trophic provisioning

Dependency can be specifically subdivided to provide clarity on the degree of the ecological benefit line-fish obtain from kelp forests, including:

-   *Facultative*: Line-fish benefit ecologically from kelp forests, but can obtain a benefit from other ecosystems
-   *Obligatory*: Line-fish strictly and only benefit ecologically from kelp forests

The Bayesian analysis framework intends on providing clarity to the interactions that occur between line-fish and kelp forests and will make use of the above the terminology. A methodology will be setup to determine the dependency ranking of line-fish, focusing specifically on trophic and habitat provisioning. The priors will be set up separately for these two ranking systems, with each focusing on factors that will determine whether fish benefit on kelp and the degree to which the benefit is received.

# Trophic Dependency Ranking:

The Bayesian analysis framework aims to determine the TDR of line-fish on kelp forests. This will follow a methodological structure, of which priors will be first be set-up. The priors selected in this section will infer knowledge on the possible interactions line-fish have with kelp forests. Due to the lack of information there are with certain fish species, uncertainties will also be taken into consideration to provide a probabilistic estimate of the interaction occurring. Priors will be set up to account for this uncertainty and will later be integrated to include the different categories to display an overall TDR of line-fish on kelp forests.

## Priors

The priors are categories that will be used to determine the possible interaction between line-fish of kelp forests. These priors will encompass factors that will touch on the boundaries and extent of fish occurrence along the coast, along with the degree of consumption of kelp or kelp associated organisms. These priors were selected based on the knowledge sourced from literature sources.

The prior represents our initial *beliefs* about a parameter's effect on the TDR estimate before observing the current data. In this context, the priors will encode our expert knowledge and assumptions about the TDR based on the answers to Questions 1-x in the Survey Questionnaire. These priors will then be updated with the data (likelihood) to produce the posterior distribution, which reflects our updated beliefs about the TDR.

Each question provides qualitative information that can be quantitatively translated into prior distributions for the TDR.

### Step 1: Assigning priors selection

The priors play a big role in shaping the posterior distributions. The type of prior---whether **Diffuse**, **Weakly Informative**, or **Informative**---reflects different levels of prior knowledge or assumptions about the parameter before seeing the data. Pay careful attention to assigning priors. A diffuse or non-informative prior is chosen to exert minimal influence on the posterior distribution and allows the data to drive the posterior in the absence of strong prior beliefs. Weakly informative priors provide some guidance without strongly biasing the posterior. They reflect a minimal level of knowledge or reasonable assumptions about the parameter's likely range without assuming high confidence in specific values. Informative priors are used when there is strong prior knowledge or beliefs about the parameter.

#### **Question 1: Trophic hierarchy**\

1.  Define Primary Categories for Trophic Level Start with three main trophic level categories:\

* 𝜃~Herbivore~: Probability of a species being strictly herbivorous. 
* 𝜃~Carnivore~: Probability of a species being strictly carnivorous. 
* 𝜃~Omnivore~: Probability of a species being omnivorous.

Define these in a Dirichlet distribution:

$$
(𝜃_{Herbivore} , 𝜃_{Carnivore} , 𝜃_{Omnivore}) ∼ Dirichlet (𝛼_{Herbivore} ,
𝛼_{Carnivore} , 𝛼_{Omnivore})
$$\

2.  Establish Proportions within the Omnivore Category\

Since an omnivore may have a mixed diet of animal and plant-based items, let's introduce two sub-parameters within the omnivore category to capture the dietary proportions: 

* 𝜙~Plant~: Proportion of the omnivore diet that is plant-based. 
* 𝜙~Animal~: Proportion of the omnivore diet that is animal-based.

These should satisfy:\

$$
𝜙_{Plant} + 𝜙_{Animal} = 1
$$\

To account for different dietary proportions, model 𝜙~Plant~ and 𝜙~Animal~ with a Dirichlet prior as well:\

$$
(𝜙_{Plant} , 𝜙_{Animal}) ∼ Dirichlet (𝛽_{Plant} , 𝛽_{Animal})
$$\

where 𝛽~Plant~ and 𝛽~Animal~ are the concentration parameters that reflect prior beliefs about the typical proportion of plant and animal components in an omnivorous diet.\

3. Integrate the Omnivore Proportions into the Overall Model\

Now,if the overall probability of an omnivorous diet is 𝜃~Omnivore~, then the expected contribution of plant and animal items to the omnivorous diet is given by:\

$$
𝜃_{Omnivore, Plant} = 𝜃_{Omnivore} × 𝜙_{Plant} 
$$\
$$
𝜃_{Omnivore, Animal} = 𝜃_{Omnivore} × 𝜙_{Animal} 
$$\

4. Define the Joint Model\

The joint model now includes both the trophic level and dietary proportions: 
• First, assign a Dirichlet distribution to the main trophic levels (𝜃~Herbivore~ ,𝜃~Carnivore~ , 𝜃~Omnivore~ ). 
• Next, conditionally on the species being omnivorous, assign a Dirichlet distribution to the proportions of plant and animal components within the omnivore category. 

The final model reflects both the probability of a species being omnivorous and the specific proportions within that dietary choice:\

$$
𝜃 =(𝜃_{Herbivore} , 𝜃_{Carnivore} , 𝜃_{Omnivore, Plant} , 𝜃_{Omnivore,Animal} ) ∼ Dirichlet(𝛼_{Herbivore}, 𝛼_{Carnivore} , 𝛼_{Omnivore} ⋅𝜙_{Plant} , 𝛼_{Omnivore} ⋅𝜙_{Animal})
$$

5. Choosing Concentration Parameters\

The concentration parameters 𝛼 and 𝛽 determine the strength of prior belief in each category. If 𝜙~Plant~ = 0.8 and 𝜙~Animal~ = 0.2, then the parameters 𝛽~Plant~ = 8 and 𝛽~Animal~ = 2 reflect moderate certainty in this distribution.Adjust 𝛼Omnivore accordingly based on the observed data or expert input.

***Process for selecting Alpha and Beta Paramaters***\

*Step 1: Understanding the Ecological Relationship*\
The expected TDR of line fish will be high for individuals that are herbivores, and smaller for line fish that are carnivores. Omnivores display a mixed diet, and will thus have an intermediate TDR.\

The TDR of omnivores will differ based on the proportion of plant and animal consumed by line fish. Omnivores with a greater proportion of plants to animals in their diet will display a greater TDR, than those with a greater proportion of animals to plants in their diet.\

*Step 2: Translating to Statistical Properties*\
For the beta distributions we will set alpha (α) and beta (β) values based on the following:

* Mean = α/(α+β)
* Mode = (α-1)/(α+β-2) [when α,β > 1]
* Variance = αβ/[(α+β)²(α+β+1)]

*Step 3: Setting Target Mean Values*\
I will select target mean values that would reflect the ecological relationships mentioned:

* Herbivore ~ 0.8 (high dependency)
* Omnivore ~ 0.55 (moderate dependency)
* Carnivores ~ 0.25 (low dependency)

These values create clear separation while remaining in the [0,1] interval appropriate for dependency measure.\

*Step 4: Determining Alpha and Beta Values*\
For each target mean, I solved for paramter combinations:

1. Herbivores (target mean = 0.8)
Using mean = α/(α+β) = 0.8

$$
\frac{α}{(α+β)} = 0.8
$$\

$$
α = 0.8 \cdot (α+β)
$$\

$$
α = 0.8α + 0.8β
$$\

$$
0.2α = 0.8β
$$\

$$
α = 4β
$$\

I selected β = 2 for moderate variance, giving α = 8\

2. Omnivores (target mean = 0.556):

* Using similar equations: α/(α+β) = 0.556

$$
\frac{α}{(α+β)} = 0.5
$$\

$$
α = 0.5 \cdot (α+β)
$$\

$$
α = 0.5α + 0.5β
$$\

$$
0.5α = 0.5β
$$\

$$
α = 1β
$$\

I selected β = 5, giving α = 5

3. Carnivores (target mean = 0.25):

* Using mean = α/(α+β) = 0.25

$$
\frac{α}{(α+β)} = 0.25
$$\

$$
α = 0.25 \cdot (α+β)
$$\

$$
α = 0.25α + 0.25β
$$\

$$
0.75α = 0.25β
$$\

$$
α = \frac{1}{3}β
$$\

I selected β = 6, giving α = 2

*Step 5: Verifying Distribution Shapes*\
For each parameter set, I verified:

* The shape matched ecological expectations
* There was appropriate separation between distributions
* Variance was reasonable (not too narrow or wide)

*Step 6: Checking Practical Interpretation*\
Finally, I confirmed that:

* The distributions aligned with ecological understanding
* The overlaps between categories reflected natural variation
* The mode and concentration were appropriate for biological variability in trophic dependencies

***Visual representation of priors***\

|**Distance (km)** | **Mean TDR** | **α** | **β** |
|------------------|--------------|-------|-------|
|**Herbivore**     | 0.8          | 8     | 2     | 
|**Omnivore**      | 0.557        | 5     | 4     |
|**Carnivore**     | 0.25         | 2     | 6     |


Plot the Beta prior distributions 

```{r}
#| echo: false
#| messages: false
#| warning: false
#| fig.cap: "Figure #: Plot of A) decay of the importance of kelp, B) total information, and C) the prior distributions as a function of depth from shore"
#| fig.align: center
#| fig-pos: "t"
#| fig.asp: 0.65



# Load packages
library(tidyverse)

# Define the parameters for each option
options <- list(
  "Herbivore" = list(alpha = 8, beta = 2),
  "Omnivore" = list(alpha = 5, beta = 5),
  "Carnivore" = list(alpha = 2, beta = 6)
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("Herbivore", "Omnivore", "Carnivore")))

# Plot the densities
class(density_long$Option)


ggplot(density_long, aes(x = TDR, y = Density, colour = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Trophic Mode Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.92),
        legend.text = element_text(size = 11))

```


#### **Question 2: Proportion of kelp/kelp associated organisms consumed** 

The proportion of stomachs that kelp or kelp associated organisms were found in can indicate the degree at which species rely on those particular food types. Species with a greater proportion of food items that are kelp/associated to kelp are more likely to have a higher TDR on kelp, while as the proportion of food items consumed decreases, the TDR on kelp is expected to decrease. The priors for this section are as follows:

***Choices: \[0 \| 0.25\| 0.5 \| 0.75 \| 1\]***

Since 1 has the greatest proportion of kelp/kelp associated organisms in the diet of line-fish, the mean will be 0.95 with a small variance. As fish with greater amounts of kelp related organisms in their diet will have a greater chance of depending on kelp, thus benefit and improve their fitness.

-   0.00 = Beta(α = 1, β = 19) mean = 0.05, va = 0.015
-   0.25 = Beta(α = 2, β = 15) me = 0.29, va = 0.026
-   0.50 = Beta(α = 10, β = 10) me = 0.5, va = 0.012
-   0.75 = Beta(α = 15, β = 5) me = 0.75, va = 0.009
-   1.00 = Beta(α = 19, β = 1) me = 0.95, va = 0.002

```{r}
# for the mean
(2)/(2+8)

# for the variance
(2*8) / ((2+8)^2 * (2+8+1))


# from claude

# Define prior parameters for different dependency levels
prior_params <- list(
  greatest_dependency = list(
    alpha = 19, beta = 1,    # Mean ~0.8, concentrated distribution
    description = "Greatest proportion of kelp/kelp-associated items"
  ),
  high_dependency = list(
    alpha = 15, beta = 5,
    description = "High proportion of kelp/kelp associated items"
  ),
  moderate_dependency = list(
    alpha = 5, beta = 5,    # Mean 0.5, moderate variance
    description = "Mixed diet with substantial kelp component"
  ),
  low_dependency = list(
    alpha = 4, beta = 10,
    description = "Low proportion of kelp items with mixed diet"
  ),
  lowest_dependency = list(
    alpha = 2, beta = 8,    # Mean ~0.2, indicates occasional kelp consumption
    description = "Lowest proportion of kelp items"
  )
)

# Function to calculate mean and variance
get_distribution_stats <- function(alpha, beta) {
  mean = alpha / (alpha + beta)
  variance = (alpha * beta) / ((alpha + beta)^2 * (alpha + beta + 1))
  return(list(mean = mean, variance = variance))
}

# Calculate metrics for each prior
for(dep_level in names(prior_params)) {
  stats <- get_distribution_stats(
    prior_params[[dep_level]]$alpha, 
    prior_params[[dep_level]]$beta
  )
  cat(sprintf(
    "%s:\n  Mean: %.2f\n  Variance: %.3f\n\n",
    dep_level, stats$mean, stats$variance
  ))
} 
```

Now lets visualize the Beta distributions for diet proportion

```{r}
#| fig.cap: "Figure 4: Plot distributions for the proportion of consumption effect on TDR"

library(tidyverse)
# Define the scaling factor 
# scale_factor <- 4.55  # Adjust this value as needed

# Define the parameters for each option
options <- list(
  "Greatest_proportion" = list(alpha = 19, beta = 1),
  "High_proportion" = list(alpha = 15, beta = 5),
  "Moderate_proportion" = list(alpha = 5, beta = 5),
  "Low_proportion" = list(alpha = 4, beta = 10),
  "None" = list(alpha = 1, beta = 19) 
)

# Create a sequence of TDR values from 0 to 1
tdr_values <- seq(0, 1, length.out = 1000)

# Initialize a data frame to hold the densities
density_data <- data.frame(TDR = tdr_values)

# Calculate the densities for each option
for (option in names(options)) {
  alpha <- options[[option]]$alpha
  beta <- options[[option]]$beta
  density_data[[option]] <- dbeta(tdr_values, alpha, beta)
}

# Convert data to long format for ggplot2
density_long <- gather(density_data, key = "Option", value = "Density", -TDR)

density_long <- density_long %>% 
  mutate(Option = factor(Option, levels = c("Greatest_proportion", 
                                            "High_proportion", 
                                            "Moderate_proportion", 
                                            "Low_proportion",
                                            "None")))



# Plot the densities
ggplot(density_long, aes(x = TDR, y = Density, color = Option)) +
#  geom_line(size = 1.3) +
  geom_area(alpha = 0.4, position = "identity", aes(fill = Option)) +
  labs(title = "Prior Distributions for TDR",
       subtitle = "Proportion of Consumption Effect",
       x = "TDR",
       y = "Density") +
  theme_minimal(base_size = 22) +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.8),
        legend.key.size = unit(0.5, 'cm'),
        legend.text = element_text(size = 11))
```
