---
title: "Zoë's Bayesian habitat and trophic evidence model"
subtitle: "A worked analysis using the strongest kelp-associated fish"
author: "AJ Smit"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: journal
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
---

# What I have done

Zoë, your central idea works. You define several lines of prior information for
habitat and trophic relationships, combine them into an integrated Beta prior,
and update that prior with the literature evidence. Habitat and trophic evidence
remain separate throughout.

I have kept that structure. This is your HDR and TDR model, rather than the
FishKelp model. I use the FishKelp analysis only to select the worked fish and
to supply a reproducible set of literature records. The response is the ordered
evidence hierarchy:

0. None
1. Co-occurrence
2. Use
3. Selection
4. Contribution
5. Dependence

The model estimates a continuous Habitat Dependency Rank (HDR) and Trophic
Dependency Rank (TDR) between zero and one. It also reports the posterior
probability assigned to each evidence level. The level probabilities are the
main result because they preserve the meaning of the evidence hierarchy.

# What needed adjustment

Your weighted Beta prior is a sensible way to combine the questionnaire
answers. The weights sum to one, so a fixed weighted pool remains a Beta
distribution. I retained your proposed weights and the exact alpha and beta
values from your worked examples. I treat the weights as uncertain rather than
fixed. Your values are the centre of a Dirichlet distribution over plausible
question weights.

The missing part was a likelihood that matches the response. The evidence
levels are ordered categories, not measured proportions. I therefore use an
ordered-probit likelihood. A record at a given level is most probable when the
unknown HDR or TDR lies within the corresponding interval. Knowledge
specificity controls the width of that observation: High-specificity records
are more precise than Medium- or Low-specificity records.

Agreement among sources supplies another form of uncertainty. The model
estimates between-source variation for each route. Several records supporting
the same level give a more focused posterior. Records spread across levels
support greater between-source variation and a wider posterior.

This calculation treats records as independent. Duplicate publications,
multiple reports from the same samples, or reviews repeating a primary result
must be grouped before fitting. Otherwise repetition can be mistaken for
independent agreement.

The simulated values in `Posteriors.qmd` cannot provide this likelihood. They
are drawn from a selected prior and fitted back to the same distribution. This
is circular and does not update the prior with observed evidence. The model
below replaces those values with coded literature records.

The questions also need a small conceptual revision before the full analysis.
Questions about geography, depth and coastal position describe access to kelp.
They can inform a prior, but they cannot establish kelp use or dependence.
Questions about occurrence, use, selection, contribution and dependence
describe evidence. Those answers must be recorded independently of the papers
used in the likelihood. Otherwise the same evidence enters the model twice.

| Questionnaire block | Recommended role in the model |
|---|---|
| Habitat and trophic Q1-Q3 | Broad prior information about access to kelp |
| Habitat Q4-Q7 | Prior information only when elicited independently; otherwise literature evidence |
| Trophic Q4 | Broad prior information about feeding mode |
| Trophic Q5-Q7 | Prior information only when elicited independently; otherwise literature evidence |

This division retains your ecological questions but separates background
traits from observations of the kelp relationship. It is the main change I
recommend for the final questionnaire.

For now, this worked analysis uses your completed HDR and TDR examples as the
priors. Your files do not name the example fish. I apply those profiles to
*Pachymetopon blochii* as a demonstration because that fish was selected by the
FishKelp analysis. Please confirm the intended example species before treating
these results as a final biological assessment.

# Set up the analysis

The paths are relative to this file, so the analysis will run from a fresh
clone of the repository.

```{r setup}
library(ggplot2)

qmd_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
analysis_dir <- dirname(qmd_path)
repo_root <- dirname(analysis_dir)

source(file.path(repo_root, "R", "zoe_hdr_tdr_model.R"))

prior_path <- file.path(repo_root, "data", "zoe_worked_prior_profiles.csv")
evidence_path <- file.path(
  repo_root,
  "data",
  "fishkelp_v7",
  "fishkelp_model_records_v7.csv"
)
fishkelp_summary_path <- file.path(
  analysis_dir,
  "posterior_outputs",
  "Pachymetopon_blochii_posterior_summary_v7.csv"
)
output_dir <- file.path(analysis_dir, "zoe_posterior_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

selected_species <- "Pachymetopon blochii"

prior_answers <- read.csv(prior_path, stringsAsFactors = FALSE)
evidence <- read.csv(
  evidence_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# The packaged source retains its original column name in position eight.
# Within this analysis we call it the evidence level.
names(evidence)[8] <- "level"

selected_evidence <- evidence[evidence$species == selected_species, ]
```

# Combine Zoë's question priors

The proposed weights are the expected relative weights. The concentration
parameter controls our confidence in them. A small concentration allows more
variation among plausible weight combinations. A large concentration keeps
the sampled weights close to the proposed values. I use 25 as a moderate
working choice and examine the effect of this decision below.

```{r integrated-priors}
habitat_answers <- prior_answers[prior_answers$route == "Habitat", ]
trophic_answers <- prior_answers[prior_answers$route == "Trophic", ]

habitat_prior <- integrate_question_priors(habitat_answers)
trophic_prior <- integrate_question_priors(trophic_answers)

weight_concentration <- 25
n_weight_draws <- 4000

habitat_prior_draws <- draw_integrated_priors(
  question_answers = habitat_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260819
)
trophic_prior_draws <- draw_integrated_priors(
  question_answers = trophic_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260820
)

integrated_priors <- rbind(
  data.frame(route = "Habitat", habitat_prior),
  data.frame(route = "Trophic", trophic_prior)
)

knitr::kable(
  integrated_priors,
  digits = 3,
  caption = "Integrated priors calculated from Zoë's worked questionnaire answers."
)

uncertain_prior_summary <- rbind(
  summarise_integrated_prior_draws(habitat_prior_draws, "Habitat"),
  summarise_integrated_prior_draws(trophic_prior_draws, "Trophic")
)

knitr::kable(
  uncertain_prior_summary,
  digits = 3,
  caption = "Variation in the prior mean after uncertainty in question weights is included."
)
```

The habitat calculation gives Beta(`r round(habitat_prior$alpha, 5)`,
`r round(habitat_prior$beta, 5)`). The trophic calculation gives
Beta(`r round(trophic_prior$alpha, 5)`,
`r round(trophic_prior$beta, 5)`). These are the central values of Zoë's priors.
The fitted model averages over the sampled alternatives. No FishKelp prior is
used.

# Prepare the literature evidence

Each paper contributes an ordered evidence level and a knowledge-specificity
class. The class controls the observation standard deviation used by the
likelihood.

```{r evidence-settings}
specificity_sd <- c(
  High = 0.10,
  Medium = 0.20,
  Low = 0.30
)

model_settings <- data.frame(
  specificity = names(specificity_sd),
  observation_sd = unname(specificity_sd)
)

knitr::kable(
  model_settings,
  caption = "Observation uncertainty used for each knowledge-specificity class."
)
```

A smaller standard deviation means that the record gives a more precise
indication of the underlying HDR or TDR. These three values are transparent
working choices. They should be checked in a sensitivity analysis before the
model is used for publication.

```{r evidence-counts}
route_names <- c(H = "Habitat", T = "Trophic")
level_names <- c(
  "None",
  "Co-occurrence",
  "Use",
  "Selection",
  "Contribution",
  "Dependence"
)

evidence_counts <- as.data.frame(
  table(
    route = factor(selected_evidence$route, levels = c("H", "T")),
    level = factor(selected_evidence$level, levels = 0:5)
  )
)
evidence_counts <- evidence_counts[evidence_counts$Freq > 0, ]
evidence_counts$route <- route_names[as.character(evidence_counts$route)]
evidence_counts$level <- paste0(
  as.integer(evidence_counts$level) - 1,
  ": ",
  level_names[as.integer(evidence_counts$level)]
)
names(evidence_counts)[3] <- "records"

knitr::kable(
  evidence_counts,
  caption = paste("Literature evidence for", selected_species)
)
```

# Fit the habitat and trophic models

The model evaluates HDR or TDR and between-source variation on a fine grid. It
averages over both question-weight uncertainty and source variation. This
is numerical integration for a small model, so Stan and Markov chain Monte
Carlo are unnecessary.

```{r fit-models}
habitat_records <- selected_evidence[selected_evidence$route == "H", ]
trophic_records <- selected_evidence[selected_evidence$route == "T", ]

habitat_fit <- fit_evidence_route(
  records = habitat_records,
  prior_draws = habitat_prior_draws,
  specificity_sd = specificity_sd
)

trophic_fit <- fit_evidence_route(
  records = trophic_records,
  prior_draws = trophic_prior_draws,
  specificity_sd = specificity_sd
)

posterior_summary <- rbind(
  summarise_evidence_route(habitat_fit, "Habitat"),
  summarise_evidence_route(trophic_fit, "Trophic")
)

level_probabilities <- rbind(
  calculate_level_probabilities(habitat_fit, "Habitat"),
  calculate_level_probabilities(trophic_fit, "Trophic")
)

write.csv(
  posterior_summary,
  file.path(output_dir, "Pachymetopon_blochii_HDR_TDR_summary.csv"),
  row.names = FALSE
)
write.csv(
  level_probabilities,
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.csv"),
  row.names = FALSE
)

knitr::kable(
  posterior_summary,
  digits = 3,
  caption = "Posterior HDR and TDR summaries for the worked fish."
)
```

The expected level is the continuous HDR or TDR multiplied by five. This places
Zoë's posterior on the 0-5 reporting scale while retaining her continuous rank.
The source-variation columns describe agreement among records. Values near zero
indicate close agreement after knowledge specificity has been considered.

# Comparison with FishKelp

FishKelp is useful here as an external benchmark. It is not used to fit Zoë's
model. Both analyses use the same fish and evidence records, but their priors
and model structures differ. Agreement would provide convergent support.
Disagreement is diagnostic and should lead us back to the questionnaire,
weights and likelihood.

```{r fishkelp-comparison}
fishkelp_summary <- read.csv(
  fishkelp_summary_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

zoe_comparison <- posterior_summary[, c(
  "route",
  "expected_level",
  "expected_level_lower_95",
  "expected_level_upper_95"
)]
names(zoe_comparison) <- c(
  "route",
  "mean",
  "lower_95",
  "upper_95"
)
zoe_comparison$model <- "Zoë HDR/TDR"

fishkelp_comparison <- data.frame(
  route = fishkelp_summary$Route,
  mean = fishkelp_summary$Mean,
  lower_95 = fishkelp_summary[["2.5%"]],
  upper_95 = fishkelp_summary[["97.5%"]],
  model = "FishKelp v7"
)

model_comparison <- rbind(zoe_comparison, fishkelp_comparison)
model_comparison <- model_comparison[
  order(model_comparison$route, model_comparison$model),
]

write.csv(
  model_comparison,
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.csv"),
  row.names = FALSE
)

knitr::kable(
  model_comparison,
  digits = 3,
  caption = "Independent model estimates on the common 0-5 evidence scale."
)
```

```{r fishkelp-comparison-plot, fig.cap="Zoë's route-specific posterior estimates compared with the FishKelp benchmark."}
#| fig-width: 8
#| fig-height: 5.5

comparison_plot <- ggplot(
  model_comparison,
  aes(x = mean, y = route, colour = model)
) +
  geom_errorbarh(
    aes(xmin = lower_95, xmax = upper_95),
    height = 0.14,
    position = position_dodge(width = 0.35),
    linewidth = 0.7
  ) +
  geom_point(
    position = position_dodge(width = 0.35),
    size = 2.8
  ) +
  scale_colour_manual(
    values = c("Zoë HDR/TDR" = "#D55E00", "FishKelp v7" = "#0072B2")
  ) +
  scale_x_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, 0.5)) +
  labs(
    title = paste("Comparison of model estimates for", selected_species),
    x = "Expected evidence level",
    y = NULL,
    colour = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.png"),
  comparison_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.pdf"),
  comparison_plot,
  width = 8,
  height = 5.5
)

comparison_plot
```

For Habitat, the two posterior means differ by only
`r round(abs(model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "FishKelp v7"]), 3)`
levels. The Habitat estimates therefore converge closely. Zoë's mean Trophic
estimate is higher by
`r round(abs(model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "FishKelp v7"]), 3)`
levels, although the 95% intervals overlap. That difference is useful. It
indicates that the trophic questions, their priors or the likelihood require
closer examination before the models can be treated as convergent.

The median between-source variation is
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Habitat"], 3)`
for Habitat and
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Trophic"], 3)`
for Trophic. The greater Trophic value reflects the wider spread of trophic
records across evidence levels and contributes to its wider posterior.

# Sensitivity to the question weights

The concentration of the Dirichlet distribution expresses how closely we hold
to the proposed weights. The values below represent broad, moderate and narrow
uncertainty around the same central weights. The biological conclusion should
not rest on one concentration value.

```{r weight-sensitivity}
weight_concentrations <- c(10, 25, 100)

fit_weight_sensitivity <- function(
  concentration,
  answers,
  records,
  route,
  seed
) {
  prior_draws <- draw_integrated_priors(
    question_answers = answers,
    weight_concentration = concentration,
    n_draws = 1500,
    seed = seed + concentration
  )
  fit <- fit_evidence_route(
    records = records,
    prior_draws = prior_draws,
    specificity_sd = specificity_sd
  )
  summary <- summarise_evidence_route(fit, route)

  data.frame(
    route = route,
    weight_concentration = concentration,
    expected_level = summary$expected_level,
    lower_95 = summary$expected_level_lower_95,
    upper_95 = summary$expected_level_upper_95
  )
}

habitat_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = habitat_answers,
    records = habitat_records,
    route = "Habitat",
    seed = 20261000
  )
)
trophic_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = trophic_answers,
    records = trophic_records,
    route = "Trophic",
    seed = 20262000
  )
)

weight_sensitivity <- rbind(
  habitat_weight_sensitivity,
  trophic_weight_sensitivity
)

write.csv(
  weight_sensitivity,
  file.path(output_dir, "Pachymetopon_blochii_weight_sensitivity.csv"),
  row.names = FALSE
)

knitr::kable(
  weight_sensitivity,
  digits = 3,
  caption = "Sensitivity of the expected evidence level to uncertainty in question weights."
)
```

# Posterior distributions

These figures follow the form of your original posterior figures. The blue
dashed line is your integrated prior. The red line is the posterior after the
literature evidence has been included. The pale vertical bands show the six
evidence levels.

```{r posterior-plot-function}
plot_posterior <- function(fit, route, rank_name) {
  level_bands <- data.frame(
    xmin = c(0, 0.10, 0.30, 0.50, 0.70, 0.90),
    xmax = c(0.10, 0.30, 0.50, 0.70, 0.90, 1),
    level = factor(0:5)
  )

  plot_data <- rbind(
    data.frame(
      rank = fit$theta,
      density = fit$prior_density,
      distribution = "Prior"
    ),
    data.frame(
      rank = fit$theta,
      density = fit$posterior_density,
      distribution = "Posterior"
    )
  )

  ggplot(plot_data, aes(x = rank, y = density)) +
    geom_rect(
      data = level_bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = level),
      inherit.aes = FALSE,
      alpha = 0.08,
      show.legend = FALSE
    ) +
    geom_line(
      aes(colour = distribution, linetype = distribution),
      linewidth = 1.05
    ) +
    scale_colour_manual(values = c(Prior = "#0072B2", Posterior = "#D55E00")) +
    scale_linetype_manual(values = c(Prior = "dashed", Posterior = "solid")) +
    scale_x_continuous(
      limits = c(0, 1),
      breaks = c(0, 0.10, 0.30, 0.50, 0.70, 0.90, 1)
    ) +
    labs(
      title = paste(route, "evidence for", selected_species),
      subtitle = "Zoë's integrated prior updated with ordered literature evidence",
      x = rank_name,
      y = "Density",
      colour = NULL,
      linetype = NULL
    ) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top")
}
```

```{r habitat-posterior, fig.cap="Prior and posterior distribution for the Habitat Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

habitat_plot <- plot_posterior(
  habitat_fit,
  route = "Habitat",
  rank_name = "Habitat Dependency Rank (HDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.png"),
  habitat_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.pdf"),
  habitat_plot,
  width = 9,
  height = 5.5
)

habitat_plot
```

```{r trophic-posterior, fig.cap="Prior and posterior distribution for the Trophic Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

trophic_plot <- plot_posterior(
  trophic_fit,
  route = "Trophic",
  rank_name = "Trophic Dependency Rank (TDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.png"),
  trophic_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.pdf"),
  trophic_plot,
  width = 9,
  height = 5.5
)

trophic_plot
```

# Probabilities for the evidence levels

The posterior distributions cross the boundaries between levels. It is more
honest to report all six probabilities than to assign a single level without
showing uncertainty.

```{r level-probability-plot, fig.cap="Posterior probability assigned to each evidence level."}
#| fig-width: 10
#| fig-height: 6

level_probabilities$evidence_level <- factor(
  level_probabilities$evidence_level,
  levels = level_names
)

level_plot <- ggplot(
  level_probabilities,
  aes(x = evidence_level, y = posterior_probability, fill = route)
) +
  geom_col(position = "dodge", width = 0.72) +
  scale_fill_manual(values = c(Habitat = "#009E73", Trophic = "#CC79A7")) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(value) paste0(round(100 * value), "%")
  ) +
  labs(
    title = paste("Evidence-level probabilities for", selected_species),
    x = "Evidence level",
    y = "Posterior probability",
    fill = "Route"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.png"),
  level_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.pdf"),
  level_plot,
  width = 10,
  height = 6
)

level_plot
```

```{r level-probability-table}
level_probability_table <- level_probabilities
level_probability_table$posterior_probability <- round(
  100 * level_probability_table$posterior_probability,
  1
)
names(level_probability_table)[4] <- "posterior probability (%)"

knitr::kable(
  level_probability_table,
  caption = "Posterior probabilities for all six evidence levels."
)
```

# How to run the full suite of fish

Zoë, the next stage is mainly data preparation. Use the following sequence for
each fish and keep habitat and trophic answers separate.

1. Copy `data/zoe_prior_profile_template.csv` for each fish. Keep
   `data/zoe_worked_prior_profiles.csv` as the completed reference example.
2. Add the scientific name in a new `species` column.
3. Complete `selected_option`, `alpha` and `beta` with the answer and parameters
   supported for that fish. Use the same central weights for every fish during
   the first full run.
4. Record the source used for every answer. Complete the questionnaire without
   consulting the literature records that will form the likelihood.
5. Draw plausible integrated priors for each route with
   `draw_integrated_priors()`.
6. Select that fish's Habitat and Trophic literature records from
   `data/fishkelp_v7/fishkelp_model_records_v7.csv`.
7. Fit the two routes separately with `fit_evidence_route()`.
8. Save the continuous HDR and TDR summaries, the six level probabilities and
   both posterior figures.
9. Repeat the analysis with narrower and wider knowledge-specificity standard
   deviations. Report conclusions that remain stable across those settings.
10. Compare each route's expected level with the independent FishKelp estimate.
    Investigate large differences; do not tune the model only to force agreement.

The complete Blamey and Bolton fish list is in
`data/fishkelp_v7/blamey_bolton_2018_fish.csv`. The packaged evidence data cover
all 30 fish, but the questionnaire answers have not yet been completed for all
species. Those answers cannot be inferred safely from a scientific name alone.

# Decisions to settle before publication

The working model runs, but six decisions still need ecological agreement.

1. Confirm the fish represented by the worked HDR and TDR examples.
2. Resolve the conflicting alpha and beta values found in the prose, tables and
   plotting code of `HDR_Priors.qmd` and `TDR_Priors.qmd`.
3. Agree on the boundaries between the six evidence levels. This analysis uses
   0.10, 0.30, 0.50, 0.70 and 0.90 as transparent working boundaries.
4. Calibrate the High, Medium and Low observation standard deviations through
   prior-predictive checks and sensitivity analysis.
5. Examine several Dirichlet concentration values for question-weight
   uncertainty. Retain the uncertainty in the final reported posteriors.
6. Identify records that share data or repeat the same primary observation.
   Count each independent line of evidence once, or model the shared source.

The model is therefore a reproducible working version of your idea. It is ready
for us to inspect and refine, but the current numerical results should not yet
be presented as final species classifications.
