---
title: "A Bayesian model for habitat and trophic evidence"
subtitle: "Working through Zoë's HDR and TDR framework for the strongest kelp-associated fish"
author: "AJ Smit"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: journal
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
---

# Aim

Zoë, I think the central idea is worth pursuing. You have defined separate sets
of questions about habitat and trophic relationships, with the intention that
the answers should inform a prior distribution and that evidence from the
literature should update it. Habitat and trophic evidence remain separate
throughout the analysis.

I have treated this as a development of your Habitat Dependency Rank (HDR) and
Trophic Dependency Rank (TDR) model, not as an implementation of FishKelp.
FishKelp serves two limited purposes here: it identifies the fish with the
strongest estimated kelp relationship, and it supplies a reproducible set of
literature records. The response follows the ordered evidence hierarchy:

0. None
1. Co-occurrence
2. Use
3. Selection
4. Contribution
5. Dependence

The model estimates continuous HDR and TDR values between zero and one and
reports the posterior probability assigned to each evidence level. I regard
these probabilities as the more useful result because they retain the meaning
of the hierarchy and show the uncertainty around any single classification.

# How I have interpreted the questions

The weighted Beta calculation gives us a prior to work with, but it is not a
formal Bayesian rule for combining independent evidence. Averaging alpha and
beta values produces a valid Beta distribution; it does not, however, follow
from updating with independent observations. I have retained the calculation
only as a way of expressing a weak context prior.

The original questions also need some revision before different people can
apply them consistently:

1. Geographic, depth and coastal overlap describe access to kelp. Some original
   priors let access imply strong use or dependence.
2. The later questions describe the same evidence that enters the likelihood.
   Using them in both places counts the evidence twice.
3. Several response labels lack numerical thresholds, so two careful extractors
   can reasonably choose different answers.
4. The prose, tables, examples and plotting code contain conflicting alpha and
   beta values.
5. Fixed weights and narrow Beta distributions imply a degree of certainty that
   we have not yet justified.

I have kept the questions but changed the role of some of them. Habitat Q1-Q3
and Trophic Q1-Q4 provide a weak context prior. The later questions are applied
to each literature source and enter through the likelihood. This separation
prevents us from using the same observation once in the prior and again in the
likelihood. The numerical priors from your worked examples remain in
`data/zoe_worked_prior_profiles.csv`, so that we do not lose the original work,
but they are not used in the analysis below.

The evidence levels are ordered categories, not measured proportions, and the
likelihood needs to respect that distinction. I use an ordered-probit
likelihood. A record at a given level is most probable when the unknown HDR or
TDR lies within the corresponding interval. The knowledge-specificity class
controls the observation error, with High-specificity records contributing
more precise information than Medium- or Low-specificity records.

The model estimates between-source variation separately for habitat and
trophic evidence. Independent records that support the same level concentrate
the posterior, whereas disagreement among records increases the estimated
variation and widens it. This gives the agreement among studies a direct role
in the uncertainty.

This calculation treats records as independent. Duplicate publications,
multiple reports from the same samples, or reviews repeating a primary result
must be grouped before fitting. Otherwise repetition can be mistaken for
independent agreement.

The simulated values in `Posteriors.qmd` do not provide such a likelihood. They
are drawn from a selected prior and fitted back to that distribution, so no
observed evidence updates the prior. The analysis below replaces those values
with literature records classified against the evidence hierarchy.

| Questionnaire block | Recommended role in the model |
|---|---|
| Habitat and trophic Q1-Q3 | Broad prior information about access to kelp |
| Habitat Q4-Q7 | Literature evidence |
| Trophic Q4 | Broad prior information about feeding mode |
| Trophic Q5-Q7 | Literature evidence |

This division retains the ecological content of the questions while separating
background traits from observations of the kelp relationship. I think this is
the main change needed before we apply the model to all species.

The revised response definitions are in
`data/zoe_prior_response_definitions.csv` and
`data/zoe_evidence_response_definitions.csv`. Each person extracting evidence
should record the response code, the supporting text or value, and its source.
“Not assessed” must remain distinct from evidence of no relationship.

# Set up the analysis

The paths are relative to this file, so the analysis will run from a fresh
clone of the repository.

```{r setup}
library(ggplot2)

qmd_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
analysis_dir <- dirname(qmd_path)
repo_root <- dirname(analysis_dir)

source(file.path(repo_root, "R", "zoe_hdr_tdr_model.R"))

prior_path <- file.path(
  repo_root,
  "data",
  "zoe_revised_prior_profile_Pachymetopon_blochii.csv"
)
prior_definitions_path <- file.path(
  repo_root,
  "data",
  "zoe_prior_response_definitions.csv"
)
evidence_definitions_path <- file.path(
  repo_root,
  "data",
  "zoe_evidence_response_definitions.csv"
)
evidence_path <- file.path(
  repo_root,
  "data",
  "fishkelp_v7",
  "fishkelp_model_records_v7.csv"
)
fishkelp_summary_path <- file.path(
  analysis_dir,
  "posterior_outputs",
  "Pachymetopon_blochii_posterior_summary_v7.csv"
)
output_dir <- file.path(analysis_dir, "zoe_posterior_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

selected_species <- "Pachymetopon blochii"

prior_answers <- read.csv(prior_path, stringsAsFactors = FALSE)
prior_definitions <- read.csv(prior_definitions_path, stringsAsFactors = FALSE)
evidence_definitions <- read.csv(
  evidence_definitions_path,
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
)
evidence <- read.csv(
  evidence_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# The packaged source retains its original column name in position eight.
# Within this analysis we call it the evidence level.
names(evidence)[8] <- "level"

selected_evidence <- evidence[evidence$species == selected_species, ]
```

# Define the responses

The context responses have target means from 0.05 to 0.20. Access to kelp can
make a relationship plausible, but it cannot by itself support use, selection,
contribution or dependence. A common concentration of eight converts each
target mean to a Beta distribution:

$$
\alpha = 8\mu, \qquad \beta = 8(1-\mu).
$$

The same rule is used for every option. These values remain expert judgements,
and we will need to assess them with prior-predictive checks. Applying one rule
does at least make the judgement explicit and avoids unrelated choices of alpha
and beta for each response.

```{r revised-prior-profile}
knitr::kable(
  prior_answers[, c(
    "route",
    "question",
    "parameter",
    "response_code",
    "selected_option",
    "alpha",
    "beta",
    "weight"
  )],
  digits = 3,
  caption = paste("Revised context-prior responses for", selected_species)
)
```

This profile remains provisional. We should calculate the geographic and depth
percentages from defined range layers before assigning a final classification;
the table specifies the required format but does not replace that work.

The evidence definitions apply a cumulative criterion: one paper contributes
the highest level supported by its observations for each route. Frequency,
dietary proportion and life stage remain useful information, but none raises
the evidence level without the observation required by the next level.

# Combine Zoë's question priors

I have taken the proposed weights as the centre of a distribution rather than
as fixed quantities. The concentration parameter controls how much the sampled
weights vary around those values: a low concentration permits more variation,
whereas a high concentration keeps them close to the proposed weights. I use 25
as a working value and examine the effect of changing it below.

```{r integrated-priors}
habitat_answers <- prior_answers[prior_answers$route == "Habitat", ]
trophic_answers <- prior_answers[prior_answers$route == "Trophic", ]

habitat_prior <- integrate_question_priors(habitat_answers)
trophic_prior <- integrate_question_priors(trophic_answers)

weight_concentration <- 25
n_weight_draws <- 4000

habitat_prior_draws <- draw_integrated_priors(
  question_answers = habitat_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260819
)
trophic_prior_draws <- draw_integrated_priors(
  question_answers = trophic_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260820
)

integrated_priors <- rbind(
  data.frame(route = "Habitat", habitat_prior),
  data.frame(route = "Trophic", trophic_prior)
)

knitr::kable(
  integrated_priors,
  digits = 3,
  caption = "Integrated weak context priors from the revised questionnaire."
)

uncertain_prior_summary <- rbind(
  summarise_integrated_prior_draws(habitat_prior_draws, "Habitat"),
  summarise_integrated_prior_draws(trophic_prior_draws, "Trophic")
)

knitr::kable(
  uncertain_prior_summary,
  digits = 3,
  caption = "Variation in the prior mean after uncertainty in question weights is included."
)
```

The habitat calculation gives Beta(`r round(habitat_prior$alpha, 5)`,
`r round(habitat_prior$beta, 5)`). The trophic calculation gives
Beta(`r round(trophic_prior$alpha, 5)`,
`r round(trophic_prior$beta, 5)`). These are the central values of the revised
priors, around which the model averages over alternative question weights. No
FishKelp prior enters this calculation.

# Prepare the literature evidence

Each paper contributes an ordered evidence level and a knowledge-specificity
class. The latter determines the observation standard deviation used in the
likelihood.

```{r evidence-settings}
specificity_sd <- c(
  High = 0.10,
  Medium = 0.20,
  Low = 0.30
)

model_settings <- data.frame(
  specificity = names(specificity_sd),
  observation_sd = unname(specificity_sd)
)

knitr::kable(
  model_settings,
  caption = "Observation uncertainty used for each knowledge-specificity class."
)
```

A smaller standard deviation gives the record more influence on the underlying
HDR or TDR. The three values used here are working choices; they need to be
examined in a sensitivity analysis and calibrated before publication.

```{r evidence-counts}
route_names <- c(H = "Habitat", T = "Trophic")
level_names <- c(
  "None",
  "Co-occurrence",
  "Use",
  "Selection",
  "Contribution",
  "Dependence"
)

evidence_counts <- as.data.frame(
  table(
    route = factor(selected_evidence$route, levels = c("H", "T")),
    level = factor(selected_evidence$level, levels = 0:5)
  )
)
evidence_counts <- evidence_counts[evidence_counts$Freq > 0, ]
evidence_counts$route <- route_names[as.character(evidence_counts$route)]
evidence_counts$level <- paste0(
  as.integer(evidence_counts$level) - 1,
  ": ",
  level_names[as.integer(evidence_counts$level)]
)
names(evidence_counts)[3] <- "records"

knitr::kable(
  evidence_counts,
  caption = paste("Literature evidence for", selected_species)
)
```

# Fit the habitat and trophic models

The model evaluates HDR or TDR and between-source variation on a fine grid,
while averaging over uncertainty in the question weights. Numerical
integration is adequate for a model of this size, so Stan and Markov chain
Monte Carlo are unnecessary here.

```{r fit-models}
habitat_records <- selected_evidence[selected_evidence$route == "H", ]
trophic_records <- selected_evidence[selected_evidence$route == "T", ]

habitat_fit <- fit_evidence_route(
  records = habitat_records,
  prior_draws = habitat_prior_draws,
  specificity_sd = specificity_sd
)

trophic_fit <- fit_evidence_route(
  records = trophic_records,
  prior_draws = trophic_prior_draws,
  specificity_sd = specificity_sd
)

posterior_summary <- rbind(
  summarise_evidence_route(habitat_fit, "Habitat"),
  summarise_evidence_route(trophic_fit, "Trophic")
)

level_probabilities <- rbind(
  calculate_level_probabilities(habitat_fit, "Habitat"),
  calculate_level_probabilities(trophic_fit, "Trophic")
)

write.csv(
  posterior_summary,
  file.path(output_dir, "Pachymetopon_blochii_HDR_TDR_summary.csv"),
  row.names = FALSE
)
write.csv(
  level_probabilities,
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.csv"),
  row.names = FALSE
)

knitr::kable(
  posterior_summary,
  digits = 3,
  caption = "Posterior HDR and TDR summaries for the worked fish."
)
```

Multiplying the continuous HDR or TDR by five places the posterior on the 0-5
reporting scale while retaining the continuous rank. The source-variation
columns describe agreement among records after allowing for knowledge
specificity; values near zero indicate closer agreement.

# Comparison with FishKelp

FishKelp provides a useful comparison, but it is not used to fit this model.
The analyses use the same fish and literature records, although their priors
and model structures differ. Similar estimates would support the consistency
of the two approaches; marked differences would direct us back to the
questions, weights and likelihood.

Because both models draw on the same literature records, this comparison is not
an independent validation. A stronger test would use species withheld during
development, new sources, or evidence extracted independently.

```{r fishkelp-comparison}
fishkelp_summary <- read.csv(
  fishkelp_summary_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

zoe_comparison <- posterior_summary[, c(
  "route",
  "expected_level",
  "expected_level_lower_95",
  "expected_level_upper_95"
)]
names(zoe_comparison) <- c(
  "route",
  "mean",
  "lower_95",
  "upper_95"
)
zoe_comparison$model <- "Zoë HDR/TDR"

fishkelp_comparison <- data.frame(
  route = fishkelp_summary$Route,
  mean = fishkelp_summary$Mean,
  lower_95 = fishkelp_summary[["2.5%"]],
  upper_95 = fishkelp_summary[["97.5%"]],
  model = "FishKelp v7"
)

model_comparison <- rbind(zoe_comparison, fishkelp_comparison)
model_comparison <- model_comparison[
  order(model_comparison$route, model_comparison$model),
]

write.csv(
  model_comparison,
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.csv"),
  row.names = FALSE
)

knitr::kable(
  model_comparison,
  digits = 3,
  caption = "Model estimates on the common 0-5 evidence scale."
)
```

```{r fishkelp-comparison-plot, fig.cap="Zoë's route-specific posterior estimates compared with the FishKelp benchmark."}
#| fig-width: 8
#| fig-height: 5.5

comparison_plot <- ggplot(
  model_comparison,
  aes(x = mean, y = route, colour = model)
) +
  geom_errorbarh(
    aes(xmin = lower_95, xmax = upper_95),
    height = 0.14,
    position = position_dodge(width = 0.35),
    linewidth = 0.7
  ) +
  geom_point(
    position = position_dodge(width = 0.35),
    size = 2.8
  ) +
  scale_colour_manual(
    values = c("Zoë HDR/TDR" = "#D55E00", "FishKelp v7" = "#0072B2")
  ) +
  scale_x_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, 0.5)) +
  labs(
    title = paste("Comparison of model estimates for", selected_species),
    x = "Expected evidence level",
    y = NULL,
    colour = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.png"),
  comparison_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.pdf"),
  comparison_plot,
  width = 8,
  height = 5.5
)

comparison_plot
```

For Habitat, the two posterior means differ by
`r round(abs(model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "FishKelp v7"]), 3)`
levels. The mean Trophic estimates differ by
`r round(abs(model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "FishKelp v7"]), 3)`
levels. The 95% intervals overlap for both routes, which suggests broad
compatibility rather than close agreement. The comparison remains approximate
because the two models define their latent quantities differently.

The median between-source variation is
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Habitat"], 3)`
for Habitat and
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Trophic"], 3)`
for Trophic. The higher Trophic value follows from the wider spread of trophic
records across evidence levels and contributes to the wider posterior.

# Sensitivity to the question weights

The concentration of the Dirichlet distribution expresses our confidence in
the proposed weights. The values below represent broad, moderate and narrow
uncertainty around the same central values. A biological conclusion should not
depend on one choice of concentration.

```{r weight-sensitivity}
weight_concentrations <- c(10, 25, 100)

fit_weight_sensitivity <- function(
  concentration,
  answers,
  records,
  route,
  seed
) {
  prior_draws <- draw_integrated_priors(
    question_answers = answers,
    weight_concentration = concentration,
    n_draws = 1500,
    seed = seed + concentration
  )
  fit <- fit_evidence_route(
    records = records,
    prior_draws = prior_draws,
    specificity_sd = specificity_sd
  )
  summary <- summarise_evidence_route(fit, route)

  data.frame(
    route = route,
    weight_concentration = concentration,
    expected_level = summary$expected_level,
    lower_95 = summary$expected_level_lower_95,
    upper_95 = summary$expected_level_upper_95
  )
}

habitat_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = habitat_answers,
    records = habitat_records,
    route = "Habitat",
    seed = 20261000
  )
)
trophic_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = trophic_answers,
    records = trophic_records,
    route = "Trophic",
    seed = 20262000
  )
)

weight_sensitivity <- rbind(
  habitat_weight_sensitivity,
  trophic_weight_sensitivity
)

write.csv(
  weight_sensitivity,
  file.path(output_dir, "Pachymetopon_blochii_weight_sensitivity.csv"),
  row.names = FALSE
)

knitr::kable(
  weight_sensitivity,
  digits = 3,
  caption = "Sensitivity of the expected evidence level to uncertainty in question weights."
)
```

The estimates change little across these concentration values because the
revised context questions have low and similar target means. The relative
question weights do not determine the result for this fish, although they may
have more influence when the context answers differ more strongly among
questions.

# Posterior distributions

These figures follow the form of your original posterior plots. The dashed blue
line shows the integrated prior, the red line shows the posterior after adding
the literature evidence, and the pale vertical bands mark the six evidence
levels.

```{r posterior-plot-function}
plot_posterior <- function(fit, route, rank_name) {
  level_bands <- data.frame(
    xmin = c(0, 0.10, 0.30, 0.50, 0.70, 0.90),
    xmax = c(0.10, 0.30, 0.50, 0.70, 0.90, 1),
    level = factor(0:5)
  )

  plot_data <- rbind(
    data.frame(
      rank = fit$theta,
      density = fit$prior_density,
      distribution = "Prior"
    ),
    data.frame(
      rank = fit$theta,
      density = fit$posterior_density,
      distribution = "Posterior"
    )
  )

  ggplot(plot_data, aes(x = rank, y = density)) +
    geom_rect(
      data = level_bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = level),
      inherit.aes = FALSE,
      alpha = 0.08,
      show.legend = FALSE
    ) +
    geom_line(
      aes(colour = distribution, linetype = distribution),
      linewidth = 1.05
    ) +
    scale_colour_manual(values = c(Prior = "#0072B2", Posterior = "#D55E00")) +
    scale_linetype_manual(values = c(Prior = "dashed", Posterior = "solid")) +
    scale_x_continuous(
      limits = c(0, 1),
      breaks = c(0, 0.10, 0.30, 0.50, 0.70, 0.90, 1)
    ) +
    labs(
      title = paste(route, "evidence for", selected_species),
      subtitle = "Zoë's integrated prior updated with ordered literature evidence",
      x = rank_name,
      y = "Density",
      colour = NULL,
      linetype = NULL
    ) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top")
}
```

```{r habitat-posterior, fig.cap="Prior and posterior distribution for the Habitat Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

habitat_plot <- plot_posterior(
  habitat_fit,
  route = "Habitat",
  rank_name = "Habitat Dependency Rank (HDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.png"),
  habitat_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.pdf"),
  habitat_plot,
  width = 9,
  height = 5.5
)

habitat_plot
```

```{r trophic-posterior, fig.cap="Prior and posterior distribution for the Trophic Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

trophic_plot <- plot_posterior(
  trophic_fit,
  route = "Trophic",
  rank_name = "Trophic Dependency Rank (TDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.png"),
  trophic_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.pdf"),
  trophic_plot,
  width = 9,
  height = 5.5
)

trophic_plot
```

# Probabilities for the evidence levels

The posterior distributions extend across the boundaries between evidence
levels. I would report all six probabilities rather than assign one level and
omit the uncertainty around it.

```{r level-probability-plot, fig.cap="Posterior probability assigned to each evidence level."}
#| fig-width: 10
#| fig-height: 6

level_probabilities$evidence_level <- factor(
  level_probabilities$evidence_level,
  levels = level_names
)

level_plot <- ggplot(
  level_probabilities,
  aes(x = evidence_level, y = posterior_probability, fill = route)
) +
  geom_col(position = "dodge", width = 0.72) +
  scale_fill_manual(values = c(Habitat = "#009E73", Trophic = "#CC79A7")) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(value) paste0(round(100 * value), "%")
  ) +
  labs(
    title = paste("Evidence-level probabilities for", selected_species),
    x = "Evidence level",
    y = "Posterior probability",
    fill = "Route"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.png"),
  level_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.pdf"),
  level_plot,
  width = 10,
  height = 6
)

level_plot
```

```{r level-probability-table}
level_probability_table <- level_probabilities
level_probability_table$posterior_probability <- round(
  100 * level_probability_table$posterior_probability,
  1
)
names(level_probability_table)[4] <- "posterior probability (%)"

knitr::kable(
  level_probability_table,
  caption = "Posterior probabilities for all six evidence levels."
)
```

# How to run the full suite of fish

Zoë, the next step is to prepare the inputs for each fish, while keeping the
habitat and trophic assessments separate.

1. Copy `data/zoe_prior_profile_template.csv` for each fish. Use
   `data/zoe_revised_prior_profile_Pachymetopon_blochii.csv` as the completed
   reference example.
2. Add the scientific name in the `species` column.
3. Select a response code from `data/zoe_prior_response_definitions.csv`. Copy
   the corresponding label, alpha and beta rather than entering new values.
4. Record the source, value and short justification used for every context
   answer. Complete these answers separately from the evidence extraction.
5. Draw plausible integrated priors for each route with
   `draw_integrated_priors()`.
6. Classify every independent literature source using
   `data/zoe_evidence_response_definitions.csv` and
   `data/zoe_evidence_extraction_template.csv`. Record “not assessed” for a
   silent route. Retain only the highest supported level per independent sample,
   fish and route, and use `source_cluster_id` to link shared samples.
7. Fit the two routes separately with `fit_evidence_route()`.
8. Save the continuous HDR and TDR summaries, the six level probabilities and
   both posterior figures.
9. Repeat the analysis with narrower and wider knowledge-specificity standard
   deviations. Report conclusions that remain stable across those settings.
10. Compare each route's expected level with the FishKelp benchmark.
    Investigate large differences; do not tune the model only to force agreement.
11. Ask a second person to classify a concealed pilot set. Compare response
    codes, highest levels and “not assessed” decisions. Resolve ambiguous
    definitions before the full extraction and report weighted agreement.

The complete Blamey and Bolton fish list is in
`data/fishkelp_v7/blamey_bolton_2018_fish.csv`. The evidence data cover all 30
fish, but the context questions have not yet been completed for every species.
We cannot infer those answers from the scientific names alone.

# Remaining flaws in the logic

Zoë, I think this version gives us a model that can be tested, but it does not
resolve several parts of the underlying logic.

1. **An evidence level is not a biological quantity.** The hierarchy describes
   the strongest type of claim supported by a study. The continuous HDR and TDR
   summarise the typical evidence across sources; they do not estimate the
   probability that a fish is biologically dependent on kelp.
2. **Averaging can obscure stronger evidence.** Many occurrence studies can
   pull the posterior down even when one well-designed experiment supports
   contribution. We need to decide whether the target is the typical
   relationship across contexts or the strongest causal claim supported by the
   evidence. These are different quantities.
3. **The literature is not a random sample.** More records are available for
   well-studied species and accessible habitats. The posterior is conditional on
   the assembled literature and its reporting biases.
4. **Agreement does not imply independence.** Papers may use the same samples,
   repeat an earlier assertion, or share a methodological bias. We must identify
   these links before agreement is allowed to increase precision.
5. **Alternative habitats are context dependent.** A fish recorded away from
   kelp can weaken a claim of obligate dependence, but it does not invalidate a
   reliable observation of kelp use elsewhere. Region, life stage and season
   may need separate estimates.
6. **The numerical calibration is provisional.** The context means, Beta
   concentration, level boundaries and specificity errors are working values,
   not empirically estimated constants.

The second point is the largest unresolved modelling choice. I suggest that a
later version should report two results: the posterior distribution of typical
HDR or TDR, and the probability that each evidential criterion has been met at
least once by independent, High-specificity evidence.

# Decisions to settle before publication

The model runs, but these decisions require ecological agreement before we use
it for a complete assessment:

1. Test the operational definitions against a concealed set of papers, revising
   them where the present wording does not resolve a genuine case.
2. Calibrate the target means and common Beta concentration through structured
   expert elicitation, prior-predictive checks and held-out species.
3. Agree on the boundaries between the six evidence levels. The present
   analysis uses 0.10, 0.30, 0.50, 0.70 and 0.90 as working boundaries.
4. Calibrate the High, Medium and Low observation standard deviations through
   prior-predictive checks and sensitivity analysis.
5. Examine several Dirichlet concentration values for question-weight
   uncertainty. Retain the uncertainty in the final reported posteriors.
6. Identify records that share data or repeat the same primary observation.
   Count each independent line of evidence once, or model the shared source.
7. Set and report an extractor-agreement threshold before the complete run.
   Preserve disagreements and their resolutions in an audit table.

This gives us a reproducible version of your idea that we can now examine and
revise. I would not yet present the numerical results as final species
classifications.
