---
title: "Zoë's Bayesian habitat and trophic evidence model"
subtitle: "A worked analysis using the strongest kelp-associated fish"
author: "AJ Smit"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: journal
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
---

# What I have done

Zoë, your central idea works. You define several lines of prior information for
habitat and trophic relationships, combine them into an integrated Beta prior,
and update that prior with the literature evidence. Habitat and trophic evidence
remain separate throughout.

I have kept that structure. This is your HDR and TDR model, rather than the
FishKelp model. I use the FishKelp analysis only to select the worked fish and
to supply a reproducible set of literature records. The response is the ordered
evidence hierarchy:

0. None
1. Co-occurrence
2. Use
3. Selection
4. Contribution
5. Dependence

The model estimates a continuous Habitat Dependency Rank (HDR) and Trophic
Dependency Rank (TDR) between zero and one. It also reports the posterior
probability assigned to each evidence level. The level probabilities are the
main result because they preserve the meaning of the evidence hierarchy.

# What needed adjustment

Your weighted Beta pool is a transparent elicitation device, but it is not a
standard Bayesian rule for combining independent evidence. Averaging alpha and
beta values produces a valid Beta distribution, but that result does not follow
from updating with independent observations. I retain the pool only to express
a weak context prior. The original ecological logic still needs revision before
different extractors can apply it consistently. Several problems matter:

1. Geographic, depth and coastal overlap describe access to kelp. Some original
   priors let access imply strong use or dependence.
2. The later questions describe the same evidence that enters the likelihood.
   Using them in both places counts the evidence twice.
3. Several response labels lack numerical thresholds, so two careful extractors
   can reasonably choose different answers.
4. The prose, tables, examples and plotting code contain conflicting alpha and
   beta values.
5. Fixed weights and precise Beta shapes give an appearance of certainty that
   has not yet been calibrated.

I have therefore kept your questions but changed their roles. Habitat Q1-Q3 and
Trophic Q1-Q4 provide a weak context prior. All later questions are applied to
each literature source and enter through the likelihood. The numerical priors
in your worked examples remain in `data/zoe_worked_prior_profiles.csv` for
traceability, but they are not used as the recommended priors below.

The missing part was a likelihood that matches the response. The evidence
levels are ordered categories, not measured proportions. I therefore use an
ordered-probit likelihood. A record at a given level is most probable when the
unknown HDR or TDR lies within the corresponding interval. Knowledge
specificity controls the width of that observation: High-specificity records
are more precise than Medium- or Low-specificity records.

Agreement among sources supplies another form of uncertainty. The model
estimates between-source variation for each route. Several records supporting
the same level give a more focused posterior. Records spread across levels
support greater between-source variation and a wider posterior.

This calculation treats records as independent. Duplicate publications,
multiple reports from the same samples, or reviews repeating a primary result
must be grouped before fitting. Otherwise repetition can be mistaken for
independent agreement.

The simulated values in `Posteriors.qmd` cannot provide this likelihood. They
are drawn from a selected prior and fitted back to the same distribution. This
is circular and does not update the prior with observed evidence. The model
below replaces those values with coded literature records.

The questions also need a small conceptual revision before the full analysis.
Questions about geography, depth and coastal position describe access to kelp.
They can inform a prior, but they cannot establish kelp use or dependence.
Questions about occurrence, use, selection, contribution and dependence
describe evidence. Those answers must be recorded independently of the papers
used in the likelihood. Otherwise the same evidence enters the model twice.

| Questionnaire block | Recommended role in the model |
|---|---|
| Habitat and trophic Q1-Q3 | Broad prior information about access to kelp |
| Habitat Q4-Q7 | Literature evidence |
| Trophic Q4 | Broad prior information about feeding mode |
| Trophic Q5-Q7 | Literature evidence |

This division retains your ecological questions but separates background
traits from observations of the kelp relationship. It is the main change I
recommend for the final questionnaire.

The revised response definitions are in `data/zoe_prior_codebook.csv` and
`data/zoe_evidence_codebook.csv`. They are deliberately explicit. Every
extractor should record a response code, the supporting text or value, and the
source. “Not assessed” is different from evidence of no relationship.

# Set up the analysis

The paths are relative to this file, so the analysis will run from a fresh
clone of the repository.

```{r setup}
library(ggplot2)

qmd_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
analysis_dir <- dirname(qmd_path)
repo_root <- dirname(analysis_dir)

source(file.path(repo_root, "R", "zoe_hdr_tdr_model.R"))

prior_path <- file.path(
  repo_root,
  "data",
  "zoe_revised_prior_profile_Pachymetopon_blochii.csv"
)
prior_codebook_path <- file.path(repo_root, "data", "zoe_prior_codebook.csv")
evidence_codebook_path <- file.path(
  repo_root,
  "data",
  "zoe_evidence_codebook.csv"
)
evidence_path <- file.path(
  repo_root,
  "data",
  "fishkelp_v7",
  "fishkelp_model_records_v7.csv"
)
fishkelp_summary_path <- file.path(
  analysis_dir,
  "posterior_outputs",
  "Pachymetopon_blochii_posterior_summary_v7.csv"
)
output_dir <- file.path(analysis_dir, "zoe_posterior_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

selected_species <- "Pachymetopon blochii"

prior_answers <- read.csv(prior_path, stringsAsFactors = FALSE)
prior_codebook <- read.csv(prior_codebook_path, stringsAsFactors = FALSE)
evidence_codebook <- read.csv(
  evidence_codebook_path,
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
)
evidence <- read.csv(
  evidence_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# The packaged source retains its original column name in position eight.
# Within this analysis we call it the evidence level.
names(evidence)[8] <- "level"

selected_evidence <- evidence[evidence$species == selected_species, ]
```

# Use the revised codebooks

The context responses have target means from 0.05 to 0.20. This is intentional:
access to kelp can make a relationship plausible, but cannot by itself support
use, selection, contribution or dependence. A common concentration of eight
converts each target mean to a Beta distribution:

$$
\alpha = 8\mu, \qquad \beta = 8(1-\mu).
$$

The same rule is used for every option. The values remain expert judgements and
must be tested through prior-predictive checks. The rule makes those judgements
visible and prevents separate, ad hoc choices of alpha and beta.

```{r revised-prior-profile}
knitr::kable(
  prior_answers[, c(
    "route",
    "question",
    "parameter",
    "response_code",
    "selected_option",
    "alpha",
    "beta",
    "weight"
  )],
  digits = 3,
  caption = paste("Revised context-prior responses for", selected_species)
)
```

This focal profile is provisional. The geographic and depth percentages should
be recalculated from explicit range layers before final classification. The
table shows the required format, not a substitute for that audit.

The evidence codebook applies a gate: one paper contributes the highest level
that its own observations support for a route. Frequency, dietary proportion
and life stage remain valuable metadata, but they do not raise a level without
the observation required by the next gate.

# Combine Zoë's question priors

The proposed weights are the expected relative weights. The concentration
parameter controls our confidence in them. A small concentration allows more
variation among plausible weight combinations. A large concentration keeps
the sampled weights close to the proposed values. I use 25 as a moderate
working choice and examine the effect of this decision below.

```{r integrated-priors}
habitat_answers <- prior_answers[prior_answers$route == "Habitat", ]
trophic_answers <- prior_answers[prior_answers$route == "Trophic", ]

habitat_prior <- integrate_question_priors(habitat_answers)
trophic_prior <- integrate_question_priors(trophic_answers)

weight_concentration <- 25
n_weight_draws <- 4000

habitat_prior_draws <- draw_integrated_priors(
  question_answers = habitat_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260819
)
trophic_prior_draws <- draw_integrated_priors(
  question_answers = trophic_answers,
  weight_concentration = weight_concentration,
  n_draws = n_weight_draws,
  seed = 20260820
)

integrated_priors <- rbind(
  data.frame(route = "Habitat", habitat_prior),
  data.frame(route = "Trophic", trophic_prior)
)

knitr::kable(
  integrated_priors,
  digits = 3,
  caption = "Integrated weak context priors from the revised questionnaire."
)

uncertain_prior_summary <- rbind(
  summarise_integrated_prior_draws(habitat_prior_draws, "Habitat"),
  summarise_integrated_prior_draws(trophic_prior_draws, "Trophic")
)

knitr::kable(
  uncertain_prior_summary,
  digits = 3,
  caption = "Variation in the prior mean after uncertainty in question weights is included."
)
```

The habitat calculation gives Beta(`r round(habitat_prior$alpha, 5)`,
`r round(habitat_prior$beta, 5)`). The trophic calculation gives
Beta(`r round(trophic_prior$alpha, 5)`,
`r round(trophic_prior$beta, 5)`). These are the central values of the revised
Zoë priors. The fitted model averages over alternative question weights. No
FishKelp prior is used.

# Prepare the literature evidence

Each paper contributes an ordered evidence level and a knowledge-specificity
class. The class controls the observation standard deviation used by the
likelihood.

```{r evidence-settings}
specificity_sd <- c(
  High = 0.10,
  Medium = 0.20,
  Low = 0.30
)

model_settings <- data.frame(
  specificity = names(specificity_sd),
  observation_sd = unname(specificity_sd)
)

knitr::kable(
  model_settings,
  caption = "Observation uncertainty used for each knowledge-specificity class."
)
```

A smaller standard deviation means that the record gives a more precise
indication of the underlying HDR or TDR. These three values are transparent
working choices. They should be checked in a sensitivity analysis before the
model is used for publication.

```{r evidence-counts}
route_names <- c(H = "Habitat", T = "Trophic")
level_names <- c(
  "None",
  "Co-occurrence",
  "Use",
  "Selection",
  "Contribution",
  "Dependence"
)

evidence_counts <- as.data.frame(
  table(
    route = factor(selected_evidence$route, levels = c("H", "T")),
    level = factor(selected_evidence$level, levels = 0:5)
  )
)
evidence_counts <- evidence_counts[evidence_counts$Freq > 0, ]
evidence_counts$route <- route_names[as.character(evidence_counts$route)]
evidence_counts$level <- paste0(
  as.integer(evidence_counts$level) - 1,
  ": ",
  level_names[as.integer(evidence_counts$level)]
)
names(evidence_counts)[3] <- "records"

knitr::kable(
  evidence_counts,
  caption = paste("Literature evidence for", selected_species)
)
```

# Fit the habitat and trophic models

The model evaluates HDR or TDR and between-source variation on a fine grid. It
averages over both question-weight uncertainty and source variation. This
is numerical integration for a small model, so Stan and Markov chain Monte
Carlo are unnecessary.

```{r fit-models}
habitat_records <- selected_evidence[selected_evidence$route == "H", ]
trophic_records <- selected_evidence[selected_evidence$route == "T", ]

habitat_fit <- fit_evidence_route(
  records = habitat_records,
  prior_draws = habitat_prior_draws,
  specificity_sd = specificity_sd
)

trophic_fit <- fit_evidence_route(
  records = trophic_records,
  prior_draws = trophic_prior_draws,
  specificity_sd = specificity_sd
)

posterior_summary <- rbind(
  summarise_evidence_route(habitat_fit, "Habitat"),
  summarise_evidence_route(trophic_fit, "Trophic")
)

level_probabilities <- rbind(
  calculate_level_probabilities(habitat_fit, "Habitat"),
  calculate_level_probabilities(trophic_fit, "Trophic")
)

write.csv(
  posterior_summary,
  file.path(output_dir, "Pachymetopon_blochii_HDR_TDR_summary.csv"),
  row.names = FALSE
)
write.csv(
  level_probabilities,
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.csv"),
  row.names = FALSE
)

knitr::kable(
  posterior_summary,
  digits = 3,
  caption = "Posterior HDR and TDR summaries for the worked fish."
)
```

The expected level is the continuous HDR or TDR multiplied by five. This places
Zoë's posterior on the 0-5 reporting scale while retaining her continuous rank.
The source-variation columns describe agreement among records. Values near zero
indicate close agreement after knowledge specificity has been considered.

# Comparison with FishKelp

FishKelp is useful here as an external benchmark. It is not used to fit Zoë's
model. Both analyses use the same fish and evidence records, but their priors
and model structures differ. Agreement would provide convergent support.
Disagreement is diagnostic and should lead us back to the questionnaire,
weights and likelihood.

This is a shared-data consistency check, not fully independent validation.
Both models use the same extracted literature records. A stronger test will use
held-out species, newly extracted sources, or an independently coded dataset.

```{r fishkelp-comparison}
fishkelp_summary <- read.csv(
  fishkelp_summary_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

zoe_comparison <- posterior_summary[, c(
  "route",
  "expected_level",
  "expected_level_lower_95",
  "expected_level_upper_95"
)]
names(zoe_comparison) <- c(
  "route",
  "mean",
  "lower_95",
  "upper_95"
)
zoe_comparison$model <- "Zoë HDR/TDR"

fishkelp_comparison <- data.frame(
  route = fishkelp_summary$Route,
  mean = fishkelp_summary$Mean,
  lower_95 = fishkelp_summary[["2.5%"]],
  upper_95 = fishkelp_summary[["97.5%"]],
  model = "FishKelp v7"
)

model_comparison <- rbind(zoe_comparison, fishkelp_comparison)
model_comparison <- model_comparison[
  order(model_comparison$route, model_comparison$model),
]

write.csv(
  model_comparison,
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.csv"),
  row.names = FALSE
)

knitr::kable(
  model_comparison,
  digits = 3,
  caption = "Model estimates on the common 0-5 evidence scale."
)
```

```{r fishkelp-comparison-plot, fig.cap="Zoë's route-specific posterior estimates compared with the FishKelp benchmark."}
#| fig-width: 8
#| fig-height: 5.5

comparison_plot <- ggplot(
  model_comparison,
  aes(x = mean, y = route, colour = model)
) +
  geom_errorbarh(
    aes(xmin = lower_95, xmax = upper_95),
    height = 0.14,
    position = position_dodge(width = 0.35),
    linewidth = 0.7
  ) +
  geom_point(
    position = position_dodge(width = 0.35),
    size = 2.8
  ) +
  scale_colour_manual(
    values = c("Zoë HDR/TDR" = "#D55E00", "FishKelp v7" = "#0072B2")
  ) +
  scale_x_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, 0.5)) +
  labs(
    title = paste("Comparison of model estimates for", selected_species),
    x = "Expected evidence level",
    y = NULL,
    colour = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.png"),
  comparison_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_model_comparison.pdf"),
  comparison_plot,
  width = 8,
  height = 5.5
)

comparison_plot
```

For Habitat, the two posterior means differ by
`r round(abs(model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Habitat" & model_comparison$model == "FishKelp v7"]), 3)`
levels. The mean Trophic estimates differ by
`r round(abs(model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "Zoë HDR/TDR"] - model_comparison$mean[model_comparison$route == "Trophic" & model_comparison$model == "FishKelp v7"]), 3)`
levels. The 95% intervals overlap for both routes. The results are broadly
compatible but not identical. The comparison scale is also approximate because
the models define their latent quantities differently.

The median between-source variation is
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Habitat"], 3)`
for Habitat and
`r round(posterior_summary$source_variation_median[posterior_summary$route == "Trophic"], 3)`
for Trophic. The greater Trophic value reflects the wider spread of trophic
records across evidence levels and contributes to its wider posterior.

# Sensitivity to the question weights

The concentration of the Dirichlet distribution expresses how closely we hold
to the proposed weights. The values below represent broad, moderate and narrow
uncertainty around the same central weights. The biological conclusion should
not rest on one concentration value.

```{r weight-sensitivity}
weight_concentrations <- c(10, 25, 100)

fit_weight_sensitivity <- function(
  concentration,
  answers,
  records,
  route,
  seed
) {
  prior_draws <- draw_integrated_priors(
    question_answers = answers,
    weight_concentration = concentration,
    n_draws = 1500,
    seed = seed + concentration
  )
  fit <- fit_evidence_route(
    records = records,
    prior_draws = prior_draws,
    specificity_sd = specificity_sd
  )
  summary <- summarise_evidence_route(fit, route)

  data.frame(
    route = route,
    weight_concentration = concentration,
    expected_level = summary$expected_level,
    lower_95 = summary$expected_level_lower_95,
    upper_95 = summary$expected_level_upper_95
  )
}

habitat_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = habitat_answers,
    records = habitat_records,
    route = "Habitat",
    seed = 20261000
  )
)
trophic_weight_sensitivity <- do.call(
  rbind,
  lapply(
    weight_concentrations,
    fit_weight_sensitivity,
    answers = trophic_answers,
    records = trophic_records,
    route = "Trophic",
    seed = 20262000
  )
)

weight_sensitivity <- rbind(
  habitat_weight_sensitivity,
  trophic_weight_sensitivity
)

write.csv(
  weight_sensitivity,
  file.path(output_dir, "Pachymetopon_blochii_weight_sensitivity.csv"),
  row.names = FALSE
)

knitr::kable(
  weight_sensitivity,
  digits = 3,
  caption = "Sensitivity of the expected evidence level to uncertainty in question weights."
)
```

The focal estimates change very little across these concentration values. That
is expected because the revised context questions all have low and similar
target means. Relative question weights are therefore not driving this worked
result. This may change for another fish with more contrasting context answers.

# Posterior distributions

These figures follow the form of your original posterior figures. The blue
dashed line is your integrated prior. The red line is the posterior after the
literature evidence has been included. The pale vertical bands show the six
evidence levels.

```{r posterior-plot-function}
plot_posterior <- function(fit, route, rank_name) {
  level_bands <- data.frame(
    xmin = c(0, 0.10, 0.30, 0.50, 0.70, 0.90),
    xmax = c(0.10, 0.30, 0.50, 0.70, 0.90, 1),
    level = factor(0:5)
  )

  plot_data <- rbind(
    data.frame(
      rank = fit$theta,
      density = fit$prior_density,
      distribution = "Prior"
    ),
    data.frame(
      rank = fit$theta,
      density = fit$posterior_density,
      distribution = "Posterior"
    )
  )

  ggplot(plot_data, aes(x = rank, y = density)) +
    geom_rect(
      data = level_bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = level),
      inherit.aes = FALSE,
      alpha = 0.08,
      show.legend = FALSE
    ) +
    geom_line(
      aes(colour = distribution, linetype = distribution),
      linewidth = 1.05
    ) +
    scale_colour_manual(values = c(Prior = "#0072B2", Posterior = "#D55E00")) +
    scale_linetype_manual(values = c(Prior = "dashed", Posterior = "solid")) +
    scale_x_continuous(
      limits = c(0, 1),
      breaks = c(0, 0.10, 0.30, 0.50, 0.70, 0.90, 1)
    ) +
    labs(
      title = paste(route, "evidence for", selected_species),
      subtitle = "Zoë's integrated prior updated with ordered literature evidence",
      x = rank_name,
      y = "Density",
      colour = NULL,
      linetype = NULL
    ) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top")
}
```

```{r habitat-posterior, fig.cap="Prior and posterior distribution for the Habitat Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

habitat_plot <- plot_posterior(
  habitat_fit,
  route = "Habitat",
  rank_name = "Habitat Dependency Rank (HDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.png"),
  habitat_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_HDR_posterior.pdf"),
  habitat_plot,
  width = 9,
  height = 5.5
)

habitat_plot
```

```{r trophic-posterior, fig.cap="Prior and posterior distribution for the Trophic Dependency Rank."}
#| fig-width: 9
#| fig-height: 5.5

trophic_plot <- plot_posterior(
  trophic_fit,
  route = "Trophic",
  rank_name = "Trophic Dependency Rank (TDR)"
)

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.png"),
  trophic_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_TDR_posterior.pdf"),
  trophic_plot,
  width = 9,
  height = 5.5
)

trophic_plot
```

# Probabilities for the evidence levels

The posterior distributions cross the boundaries between levels. It is more
honest to report all six probabilities than to assign a single level without
showing uncertainty.

```{r level-probability-plot, fig.cap="Posterior probability assigned to each evidence level."}
#| fig-width: 10
#| fig-height: 6

level_probabilities$evidence_level <- factor(
  level_probabilities$evidence_level,
  levels = level_names
)

level_plot <- ggplot(
  level_probabilities,
  aes(x = evidence_level, y = posterior_probability, fill = route)
) +
  geom_col(position = "dodge", width = 0.72) +
  scale_fill_manual(values = c(Habitat = "#009E73", Trophic = "#CC79A7")) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(value) paste0(round(100 * value), "%")
  ) +
  labs(
    title = paste("Evidence-level probabilities for", selected_species),
    x = "Evidence level",
    y = "Posterior probability",
    fill = "Route"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.png"),
  level_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(output_dir, "Pachymetopon_blochii_level_probabilities.pdf"),
  level_plot,
  width = 10,
  height = 6
)

level_plot
```

```{r level-probability-table}
level_probability_table <- level_probabilities
level_probability_table$posterior_probability <- round(
  100 * level_probability_table$posterior_probability,
  1
)
names(level_probability_table)[4] <- "posterior probability (%)"

knitr::kable(
  level_probability_table,
  caption = "Posterior probabilities for all six evidence levels."
)
```

# How to run the full suite of fish

Zoë, the next stage is mainly data preparation. Use the following sequence for
each fish and keep habitat and trophic answers separate.

1. Copy `data/zoe_prior_profile_template.csv` for each fish. Use
   `data/zoe_revised_prior_profile_Pachymetopon_blochii.csv` as the completed
   reference example.
2. Add the scientific name in the `species` column.
3. Select only a response code defined in `data/zoe_prior_codebook.csv`. Copy
   the label, alpha and beta from that row rather than entering new values.
4. Record the source, value and short justification used for every context
   answer. Complete these answers separately from the evidence extraction.
5. Draw plausible integrated priors for each route with
   `draw_integrated_priors()`.
6. Code every independent literature source with
   `data/zoe_evidence_codebook.csv` and
   `data/zoe_evidence_extraction_template.csv`. Record “not assessed” for a
   silent route. Retain only the highest supported level per independent sample,
   fish and route, and use `source_cluster_id` to link shared samples.
7. Fit the two routes separately with `fit_evidence_route()`.
8. Save the continuous HDR and TDR summaries, the six level probabilities and
   both posterior figures.
9. Repeat the analysis with narrower and wider knowledge-specificity standard
   deviations. Report conclusions that remain stable across those settings.
10. Compare each route's expected level with the FishKelp benchmark.
    Investigate large differences; do not tune the model only to force agreement.
11. Have a second extractor code a concealed pilot set. Compare response codes,
    highest levels and “not assessed” decisions. Revise ambiguous definitions
    before the full extraction and report weighted agreement.

The complete Blamey and Bolton fish list is in
`data/fishkelp_v7/blamey_bolton_2018_fish.csv`. The packaged evidence data cover
all 30 fish, but the questionnaire answers have not yet been completed for all
species. Those answers cannot be inferred safely from a scientific name alone.

# Remaining flaws in the logic

Zoë, I think the revised framework is coherent enough to test, but several
limitations remain important.

1. **An evidence level is not a biological amount.** The hierarchy describes
   the strongest type of claim supported by a study. The continuous HDR and TDR
   summarise the typical evidence across sources. They are not the probability
   that a fish is biologically dependent on kelp.
2. **Averaging can hide decisive evidence.** Many occurrence papers can pull a
   posterior down even when one strong experiment supports contribution. We
   must decide whether the target is the typical relationship across contexts
   or the highest causal claim supported. Those are different questions.
3. **The literature is not a random sample.** Well-studied species and accessible
   habitats generate more records. The posterior is conditional on the assembled
   corpus and its reporting biases.
4. **Agreement is not automatically independence.** Papers can use the same
   samples, repeat the same assertion or share the same method bias. Source
   clusters must be recorded before agreement is allowed to increase precision.
5. **Observed alternatives are context dependent.** A fish recorded away from
   kelp can weaken a claim of obligate dependence, but it does not erase a valid
   observation of kelp use elsewhere. Region, life stage and season may need
   separate estimates.
6. **The numerical calibration is provisional.** The context means, Beta
   concentration, level boundaries and specificity errors are transparent
   working values, not empirically established constants.

The second point is the largest unresolved modelling choice. A useful next
version may report two outputs: the posterior distribution of typical HDR or
TDR, and the probability that each causal gate has been passed at least once by
independent high-specificity evidence.

# Decisions to settle before publication

The working model runs, but seven decisions still need ecological agreement.

1. Review the operational definitions against a concealed set of real papers.
   Revise a definition only when the current wording cannot resolve the case.
2. Calibrate the target means and common Beta concentration through structured
   expert elicitation, prior-predictive checks and held-out species.
3. Agree on the boundaries between the six evidence levels. This analysis uses
   0.10, 0.30, 0.50, 0.70 and 0.90 as transparent working boundaries.
4. Calibrate the High, Medium and Low observation standard deviations through
   prior-predictive checks and sensitivity analysis.
5. Examine several Dirichlet concentration values for question-weight
   uncertainty. Retain the uncertainty in the final reported posteriors.
6. Identify records that share data or repeat the same primary observation.
   Count each independent line of evidence once, or model the shared source.
7. Set and report an extractor-agreement threshold before the complete run.
   Preserve disagreements and their resolutions in an audit table.

The model is therefore a reproducible working version of your idea. It is ready
for us to inspect and refine, but the current numerical results should not yet
be presented as final species classifications.
